\documentclass{beamer}

\title{Multiple Regression}
\author{Jeffrey B. Arnold}
\date{January 28, 2016}

\usepackage{tikz}
\usepackage{amsmath}

\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\cor}{Cor}
\DeclareMathOperator{\mean}{mean}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\MSE}{MSE}
\DeclareMathOperator{\Bias}{Bias}
\DeclareMathOperator{\SE}{se}
\DeclareMathOperator{\SD}{sd}
\DeclareMathOperator{\argmin}{argmin}
\DeclareMathOperator{\argmax}{argmax}

\usepackage{Sweave}
\begin{document}
\input{multivariate_regression-concordance}

\begin{frame}
  \maketitle{}
\end{frame}

\section{Estimating Multiple Regression}

\begin{frame}
  \frametitle{Multiple Regression}
  
  \begin{itemize}
  \item What is (are) the linear regression for multiple linear regression?
  \item What is the sample statistic used to estimate multiple linear regression?
  \item Linear regression finds the values of $\hat\beta$ that minimize what?
  \item What is the mean value of the errors in linear regression? 
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Multiple Linear Regression}
  
  \begin{itemize}
  \item The population regression equation* is:
  \begin{equation*}
  Y = \beta_0 + \sum_{k = 1}^{K} \beta_k X_k + \epsilon
  \end{equation*}
  where $\E(\epsilon) = 0$.
  \item The sample regression equation is:
  \begin{equation*}
  y_i = \hat\beta_0 + \sum_{k = 1}^K  \beta_k x_{k, i} + \hat\epsilon_i
  \end{equation*}
  \item Linear regression finds the values of $\hat\beta$ that minimize the sum of squared errors
  \begin{equation*}
  \hat\beta = \argmin_{\tilde\beta} \sum_{i = 1}^n (y_i - \beta_0 - \sum_{k = 1}^K \beta_k x_i)
  \end{equation*}
  \item The solution that ordinary least squares finds has $\mean(\hat\epsilon) = 0$
  \end{itemize}
  
\end{frame}

\begin{frame}
\frametitle{Multiple Linear Regression}
\begin{itemize}
\item You want to estimate the effect of race on approval for Donald Trump (measured on a thermometer 1-100)? You sample 1000 whites? You cannot estimate the regression 
  \begin{equation*}
  \text{thermometer} = \beta_0 + \beta_1 \text{race} + \epsilon
  \end{equation*}
  Why not?
\end{itemize}
\end{frame}

\section{Anatomy of OLS Coefficients}

\begin{frame}
\frametitle{Estimating Multiple Linear Regression}

  In a bivariate linear regression the coefficients are
  \begin{equation*}
  y_i = \hat\beta_1 + \hat\beta_1 x_i + \hat\epsilon_i
  \end{equation*}
  where
  \begin{align*}
  \hat\beta_1 &= \frac{\sum (x_i - \bar{x}) (y_i - \bar{y})}{\sum (x_i - \bar{x})^2} = \frac{\cov(x, y)}{\var{x, y}} \\
  \hat\beta_0 &= \bar{y} - \hat{\beta}_1 \bar{x}
  \end{align*}
  
  What about in multiple linear regression?

\end{frame}

\begin{frame}
\frametitle{Anatomy of Multiple Linear Regression Coefficients}

See exercise

\end{frame}

\begin{frame}
\frametitle{Anatomy of Multiple Linear Regression Coefficients}

The regression coefficients for variable $k$ is
\begin{equation*}
\hat{\beta}_k = \frac{\cov(\tilde{\epsilon}_1, \hat{\eta}_k)}{\var(\tilde{\epsilon}_k)}
\end{equation*}
where
\begin{itemize}
\item $\tilde{\eta}$ are the residuals from the regression of $y$ on other $x$.
\item $\tilde{\epsilon}$ are the residuals from the regression of $x_k$ on other $x$.
\end{itemize}


\end{frame}

\section{Omitted Varible Bias}

\begin{frame}
\frametitle{Omitted Variable Bias I}

Suppose the true model is 
\begin{equation*}
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \epsilon
\end{equation*}
However, we estimate the model,
\begin{equation*}
y_i = \hat\beta_0 \hat\beta_1 x_1 + \hat\epsilon_i
\end{equation*}
What is the relationship between $\hat\beta_1$ and $\beta_1$?
\end{frame}

\begin{frame}
\frametitle{Omitted Variable Bias II}

The coefficient in the regression with an omitted $x_2$ is,
\begin{equation*}
\hat\beta_1 = \beta_1 + \beta_2 \frac{\cov(x, z)}{\var{x}}
\end{equation*}

\begin{itemize}
\item What is the bias of $\hat\beta_1$?
\item For what values of $\beta_2$ and $\cov(x, z)$ is $\hat\beta_1$ unbiased? Interpret them.
\end{itemize}
\end{frame}


\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

%  LocalWords:  DGP dgp SMISS
