---
title: "Examples from Class"
author: "Jeffrey Arnold"
date: "February 18, 2016"
---

```{r echo=FALSE,results='hide'}
source("init.R")
```

# 2016 Nevada Primaries

Consider this CNN poll of the Nevada Democratic and Republican Presidential Caucuses taken on Feb 10-12, 2016: http://elections.huffingtonpost.com/pollster/polls/cnn-23764.
The full poll results are here: http://i2.cdn.turner.com/cnn/2016/images/02/17/topnv1.pdf

*Calculate the standard error and 95% confidence interval of Donald Trump's proportion of support?*

The CNN poll has a sample of 245 likely voters (Republican), and the proportion who will vote for Trump as 45%.

The 95% confidence interval is about 38.8 to 51.2% 
```{r}
p <- 0.45
n <- 245
p + c(-1, 1) * 1.96 * sqrt(p * (1 - p) / n)
```

*Calculate the standard error and 95% confidence interval of Bernie Sanders proportion of support?*

The CNN poll has a sample of 282 likely voters (Democratic), and the proportion who answered that they will vote for Sanders is 47%.

The 95% confidence interval is 41.2--52.8%
```{r}
p <- 0.47
n <- 282
p + c(-1, 1) * 1.96 * sqrt(p * (1 - p) / n)
```

*Can you calculate the confidence interval of the difference between Ted Cruz and Marco Rubio's support? What assumption is violated?*

No. These are not independent populations.

*Look at the most recent values of Huffpost Pollster Trend for the 2016 National Republican Primary support by Republicans for the Republican candidates. Use a Chi-squared test to compare the poll-results in Nevada to the national trend. Are they different than the national trend?*

Pollster uses a model that approximately averages the polls.  On [Feb 17](http://elections.huffingtonpost.com/pollster/2016-national-gop-primary), it HuffPost Pollster's model estimated the support for the candidates at
```{r}
national = c("Donald Trump" = 0.378,
                   "Ted Cruz" = .181,
                   "Marco Rubio" = .149,
                   "John Kasich" = .078,
                   "Ben Carson" = .083,
                   "Jeb Bush" = .063,
                   "Other" = .068)
```
The values observed in the CNN poll for Nevada are,
```{r}
n <- 245
poll_results = c("Donald Trump" = 0.45,
                   "Ted Cruz" = .17,
                   "Marco Rubio" = .19,
                   "John Kasich" = .05,
                   "Ben Carson" = .07,
                   "Jeb Bush" = .01,
                   "Other" = .06)

chisq.test(poll_results, p = national)
```
The test does not reject that the Nevada primary opinion is different than national opinion, but the small sample size makes the Chi-squared test inappropriate.


*Is support for Hillary Clinton and Bernie Sanders independent of age? See p. 33 of the full poll results. Use a Chi-squared test of independence.*


*Is support for Hillary Clinton different between men and women? Use a confidence interval and a hypothesis test for a difference of proportions.*


# Interpreting Difference in Means

## Statistically Significant and Not Statistically Significant

Consider two samples of size 100 from independent populations.
Sample 1 has a mean of $\bar{x}_1 = 0$, standard deviation of $s_1 = 10$.
Sample 2 has a mean of $\bar{x}_2 = 2.3$, and standard deviation of $s_2 = 10$.

- For sample 1, calculate its 95% confidence interval. Can you reject $H_0: \mu_1 = 0$ at the 5% significance level (assume a two-sided test)?
- Do the same for sample 2? 
- Calculate the confidence interval for the difference $\mu_1 - \mu_2$? Can you reject $H_0: \mu_1 = \mu_2$? 

```{r}
x_bar_1 <- 0
s_1 <- 10
n_1 <- 100
se_1 <- s_1 / sqrt(n_1)

x_bar_2 <- 2.3
s_2 <- s_1
n_2 <- n_1
se_2 <- s_2 / sqrt(n_2)
```

The 95% confidence interval of $\mu_1$ includes 0, obviously, since $\bar{x}_1 = 0$. The hypothesis test implied by the confidence interval does not reject that $\mu_1 = 0$.
```{r}
x_bar_1 + c(-1, 1) * 1.96 * se_1
```
The 95% confidence interval of $\mu_2$. The hypothesis test implied by the confidence interval does reject that $\mu_2 = 0$, and accepts the alternative hypothesis that $\mu_2 \neq 0$.
```{r}
x_bar_2 + c(-1, 1) * 1.96 * se_2
```

Although our tests show that $\mu_2 \neq 0$ and cannot reject $\mu_1 = 0$, however the data cannot reject that $\mu_1 = \mu_2$.
The 95% confidence interval of $\mu_1 - \mu_2$ is
```{r}
se_diff <- sqrt(se_1 ^ 2 + se_2 ^ 2)
(x_bar_1 - x_bar_2) + c(-1, 1) * 1.96 * se_diff
```
The confidence interval includes 0, so the data cannot reject the null hypothesis that $\mu_1 = \mu_2$.

Thus although $\mu_2$ is statistically significant, and $\mu_1$ is not statistically significant, this does not imply that the difference between them is statistically significant.
This is unfortunately a common interpretative mistake in applied work. 
For more, see

> Gelman, Andrew, and Hal Stern. 2006. ``The Difference Between `Significant' and `Not Significant' is not Itself Statistically Significant.'' *The American Statistician.* <https://dx.doi.org/10.1198/000313006X152649>.

## Non Overlapping Confidence Intervals

Consider two samples of size 100 from independent populations.
Sample 1 has a mean of $\bar{x}_1 = 0$, standard deviation of $s_1 = 10$.
Sample 2 has a mean of $\bar{x}_2 = 4$, and standard deviation of $s_2 = 10$.

- Calculate the 95% confidence intervals of $\mu_1$, and $\mu_2$?
- Do the confidence intervals of the mean for sample 1 and sample 2 overlap? From this can you conclude that you can reject $H_0: \mu_1 = \mu_2$ at the the 5% significance level?
- Calculate the confidence interval of $\mu_1 - \mu_2$. From this can you conclude that you can reject $H_0: \mu_1 = \mu_2$?

```{r}
x_bar_1 <- 0
s_1 <- 10
n_1 <- 100
se_1 <- s_1 / sqrt(n_1)

x_bar_2 <- 4
s_2 <- s_1
n_2 <- n_1
se_2 <- s_2 / sqrt(n_2)
```

The 95% confidence interval of $\mu_1$ is
```{r}
x_bar_1 + c(-1, 1) * 1.96 * se_1
```
The 95% confidence interval of $\mu_2$ is
```{r}
x_bar_2 + c(-1, 1) * 1.96 * se_2
```
They do not overlap.

The 95% confidence interval of $\mu_1 - \mu_2$ does not include 0, so the data rejects that $\mu_1 = \mu_2$ and accepts the alternative hypothesis that $\mu_1 \neq \mu_2$,
```{r}
se_diff <- sqrt(se_1 ^ 2 + se_2 ^ 2)
(x_bar_1 - x_bar_2) + c(-1, 1) * 1.96 * se_diff
```


## Overlapping Intervals

Consider two samples of size 100 from independent populations.
Sample 1 has a mean of $\bar{x}_1 = 0$, standard deviation of $s_1 = 10$.
Sample 2 has a mean of $\bar{x}_2 = 3$, and standard deviation of $s_2 = 10$.

- Calculate the 95% confidence intervals of $\mu_1$, and $\mu_2$?
- Do the confidence intervals of the mean for sample 1 and sample 2 overlap? From this can you conclude that you can reject $H_0: \mu_1 = \mu_2$ at the the 5% significance level?
- Calculate the confidence interval of $\mu_1 - \mu_2$. From this can you conclude that you can reject $H_0: \mu_1 = \mu_2$?

```{r}
x_bar_1 <- 0
s_1 <- 10
n_1 <- 100
se_1 <- s_1 / sqrt(n_1)

x_bar_2 <- 3
s_2 <- s_1
n_2 <- n_1
se_2 <- s_2 / sqrt(n_2)
```

The 95% confidence interval of $\mu_1$:
```{r}
x_bar_1 + c(-1, 1) * 1.96 * se_1
```
The 95% confidence interval of $\mu_2$:
```{r}
x_bar_2 + c(-1, 1) * 1.96 * se_2
```

However, the 95% confidence interval of $\mu_1 - \mu_2$ does not contain 0, so the data rejects the hypothesis that $\mu_1 = \mu_2$ and accepts the alternative hypothesis that $\mu_1 \neq \mu_2$,
```{r}
se_diff <- sqrt(se_1 ^ 2 + se_2 ^ 2)
(x_bar_1 - x_bar_2) + c(-1, 1) * 1.96 * se_diff
```
So although the confidence intervals of two groups overlaps, the difference between these groups may be statistically significant.
In order to check whether the difference between the groups is statistically significant is to test the difference.

Why do the confidence intervals of the means of each group and the confidence interval of the difference of means produce different results.
Confidence intervals of the means of two groups will not overlap when
$$
\bar{x}_1 - \bar{x_2} > z^* (se_1 + se_2) .
$$
The confidence interval of the difference in means of two groups will not include zero when
$$
\bar{x}_1 - \bar{x_2} > z^* \sqrt{se_1^2 + se_2^2} .
$$
Since the sum of the standard errors of the means of two samples is always greater than the standard error of difference in means,
$$
(se_1 + se_2) > \sqrt{se_1^2 + se_2^2}
$$
Thus there are cases in which the confidence intervals of two means can overlap but the confidence interval of the difference in means does not include zero.

We observed this in the example, 
```{r}
se_1 + se_2
se_diff
```

Thus, if two confidence intervals are **not** overlapping, then the difference will also be significant.
But, if the two means are overlapping, the difference may or may not be significant.

For more see this [handout](https://www.cscu.cornell.edu/news/statnews/Stnews73insert.pdf)

