---
title: "Commons Errors in R and what to do about them"
author: "Jeffrey Arnold"
date: "January 19, 2016"
output: html_document
---

```{r results='hide',ask=FALSE}
source("init.R")
```


```
(1 + (2 * 4)
```

```
1 + (2 * 3))
```

```
print("Hello, World!)
```

```
print(Hello, World!)
```

```
print(Hello)
```

```
print("R is awesome!"")
```

```
blah(2)
```

```
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) +
  geom_point()
```

```
2 + blah
```

```
2 + "blah"
```

<!--
- `Could not find function`. Function does not exist, or you did not load a package with that function.
- `Object '???' not found`
- `Error in if` Not a logical value
- `cannot open the connection`, `no such file or directory`: You've given R the wrong path. Check the URL or file path.
- `no applicable method for '???' applied to an object of class "???"`, `no method for coercing ...`, `no slot of name "group" for this object of class "???"`
- `subscript out of bounds` Access an element of a vector or list that is not there
- `Error in ...: replacement has ...`
- `built under R package`
- messages in ggplot2
-->


```
iris
colnames(iris)
iris[ , "Species"]
iris[ , Species]
iris[["Species"]]
iris[[Species]]
iris[["blahblah"]]
iris[ , "blahblah"]

iris[["Species"]] <- NULL
iris[["Species"]]
```
 
```
read.csv("this_file_probably_does_not_exist.csv")
```

Load **dplyr**, then load the package **MASS**.
```{r}
library(dplyr)
library(MASS)
```
What are the messages shown when you load them?
Now try using the dataset `cars` in **ggplot2**,
```
data(mpg, package = "ggplot2")
```
Select the variables `manufacturer`, `model`, and `year` from `mpg`.
```
select(mpg, manufacturer, model, year)
```
What's going on? 

`r solution_start()`
Since you loaded `MASS` second, **R** uses the function `select` from that package, not
from `dplyr`. You can reference a specific function in a package using `::`, e.g. `dplyr::select` for the `select` function in **dplyr**, and `MASS::select` for the select function in **MASS**.
```
dplyr::select(mpg, manufacturer, model, year)
```
`r solution_end()`

This problem can also occur if you define functions or variables that have the same
names as functions and variables from packages or in base R.
Define a new function named `ggplot`,
```
ggplot <- function(x) print(x)
```
Try creating a plot with the `ggplot` function
```
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) +
  geom_point()
```
However, you can still create a plot with 
```
ggplot2::ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) +
  geom_point()
```


# References

- http://www.burns-stat.com/documents/tutorials/impatient-r/#errors
- http://varianceexplained.org/pages/errors.html
- http://blog.revolutionanalytics.com/2015/03/the-most-common-r-error-messages.html
- https://github.com/noamross/zero-dependency-problems/blob/master/misc/stack-overflow-common-r-errors.md
