---
title: "Covariance and Correlation"
author: "Jeffrey Arnold"
date: "February 25, 2016"
output: html_document
---

```{r message=FALSE}
library("dplyr")
library("tidyr")
library("broom") 
library("ggplot2")
library("uwpols501")
```


## Covariance

The sample covariance of $x$ and $y$ is,
$$
\Cov(x, y) = \frac{1}{n - 1} \sum_i (x_i - \bar{x}) (y_i - \bar{y})
$$

There is also corresponding parameter for the population defined in terms of the expectation of random variables.
The covariance of random variables $X$ and $Y$ are,
$$
\Cov(X, Y) = (X - \E(X)) (Y - \E(Y))
$$

- The covariance of a variable with itself is its variance,
  $$
  Cov(x, x) = \frac{1}{n - 1} \sum_i (x_i - \bar{x}) (x_i - \bar{x}) = Var(x)
  $$
- Like variance, the covariance includes the term $1 / (n - 1)$. This is because dividing by $n - 1$ provides an unbiased estimate of the population covariance.
- The range of covariance is $(-\infty, +\infty)$.
- Covarariance does not change when a constant is added to either (or both) variables,
  $$
  \begin{aligned}[t]
  Cov(x + a, y) = \frac{1}{n - 1} \sum_i ((x_i + a) - \overline{(x + a)} - b) (y_i - \bar{y}) \\
  = Cov(x + a, y) = \frac{1}{n - 1} \sum_i ((x_i + a) - \bar{x} - a) (y_i - \bar{y}) \\
  = Cov(x + a, y) = \frac{1}{n - 1} \sum_i (x_i + \bar{x}) (y_i - \bar{y}) \\
  = Cov(x, y)
  \end{aligned}[t]  
  $$
- Covariance does change when either (or both) variables is multiplied by a constant. This implies that the scale of the variables affects the covariance.
  $$
  \begin{aligned}[t]
  Cov(a x, y) = \frac{1}{n - 1} \sum_i (a x_i - \overline{(a x)) (y_i - \bar{y}) \\
  = Cov(a x, y) = \frac{1}{n - 1} \sum_i (ax_i - a\bar{x}) (y_i - \bar{y}) \\
  = Cov(a x, y) = \frac{a}{n - 1} \sum_i (x_i + \bar{x}) (y_i - \bar{y})
  = a Cov(x, y)
  \end{aligned}[t]
  $$
  

## Correlation

The sample covariance of $x$ and $y$ is the covariance divided by the 
$$
\Cor(x, y) = \Cov(x, y) / (\sd(x) \sd(y)) 
$$

- The correlation of a variable with itself is equal to one,
  $$
  Cor(x, x) = \Cov(x, x) / (\sd(x) \sd(x)) = \Var(x) / \Var(x) = 1
  $$
- The range of covariance is $(-1, 1)$.
- Correlation does not change when a constant is added to either (or both) variables,
  $$
  \begin{aligned}[t]
  \Cor(x + a, y) = \frac{\Cov(x + a, y)}{\sd(x + a) \sd(y)} = \frac{\Cov(x, y) }{ \sd(x) \sd(y)} = \Cor(x, y)
  \end{aligned}[t]  
  $$
- Unlike covariance, correlation does not change when either (or both) variables is multiplied by a constant. Thus, the correlation of two variables, is not affected by the scale of those variables.
  $$
  \begin{aligned}[t]
  \Cor(ax, y) = \frac{\Cov(ax, y)}{\sd(ax) \sd(y)} = \frac{a \Cov(x, y) }{a \sd(x) \sd(y)} = \Cor(x, y)
  \end{aligned}[t]
  $$
  
### Example

What is the relationship between the log effective number of parties and redistribution?

```{r}
data("iver", package = "uwpols501")
```


```{r}
iver_cov_table_1 <- 
  iver %>%
    mutate(ln_enp = log(enp),
           ln_enp_diff = ln_enp - mean(ln_enp),
           povred_diff = povred - mean(povred),
           povred_diff_ln_enp_diff = ln_enp_diff * povred_diff) %>%
    select(cty, povred, ln_enp, povred_diff,ln_enp_diff, povred_diff_ln_enp_diff)
iver_cov_table_1 %>%
  knitr::kable()
```
The covariance is the sum of the product of the differences between the variables and their means,
```{r}
summarize(iver_cov_table_1, sum(povred_diff_ln_enp_diff) / (n() - 1))
```
We can confirm this using the `cov` function
```{r}
cov(iver$povred, log(iver$enp))
```
Note that the covariance does not depend on the order of the variables,
```{r}
cov(log(iver$enp), iver$povred)
```

The correlation divides the covariance by the product of the standard deviations of each variable,
```{r}
cov(iver$povred, log(iver$enp)) / (sd(iver$povred) * sd(log(iver$enp)))
```
This agrees with the correlation as produced by `cor`,
```{r}
cor(iver$povred, log(iver$enp))
```


# 


## Anscombe Quartet

The Anscombe quartet is a data frame included in 
```{r}
data("anscombe", package = "datasets")
```
```{r}
anscombe_df <- 
  as_data_frame(anscombe) %>%
    mutate(id = row_number()) %>%
    gather(variable_, value, -id) %>%
    separate(variable_, c("variable", "dataset"), sep = 1) %>%
    spread(variable, value)
```

What are the means, standard deviations, and correlations of $x$ and $y$ in each dataset,
```{r}
anscombe_df %>%
  group_by(dataset) %>%
  summarize(mean_x = mean(x),
            mean_y = mean(y),
            sd_x = sd(x),
            sd_y = sd(y),
            cor_xy = cor(x, y),
            cov_xy = cov(x, y))
```
These datasets have the same means for $x$ and $y$, and approximately the same standard deviations of $x$ and $y$, and the approximately same correlation and covariance between $x$ and $y$. 
If we run linear regression on each of these, they have the same intercepts and slopes.
```{r}
anscombe_df %>%
  group_by(dataset) %>%
  do(tidy(lm(y ~ x, data = .))) %>%
  select(term, estimate) %>%
  spread(term, estimate)
```

However, if we plot these datasets, although they have the same linear relationship between $x$ and $y$, they look very different.
```{r}
ggplot(anscombe_df, aes(x = x, y = y)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~dataset)
```

Note that correlation and covariance only measure the strength and direction of the linear relationship between two variables. 
It does not say anything about the slope of that linear relationship.
Nor does it capture non-linear relationships between variables.
The following figure illustrates these points.
The figure plots several sets of x,y points and the correlation coefficient between those varaibles.

![Several sets of (x, y) points, with the Pearson correlation coefficient of x and y for each set](images/Correlation_examples2.svg)

By <a href="//commons.wikimedia.org/w/index.php?title=User:DenisBoigelot&amp;action=edit&amp;redlink=1" class="new" title="User:DenisBoigelot (page does not exist)">DenisBoigelot</a>, original uploader was <a href="//en.wikipedia.org/wiki/User:Imagecreator" class="extiw" title="en:User:Imagecreator">Imagecreator</a> - <span class="int-own-work" lang="en">Own work</span>, original uploader was <a href="//en.wikipedia.org/wiki/User:Imagecreator" class="extiw" title="en:User:Imagecreator">Imagecreator</a>, <a href="http://creativecommons.org/publicdomain/zero/1.0/deed.en" title="Creative Commons Zero, Public Domain Dedication">CC0</a>, https://commons.wikimedia.org/w/index.php?curid=15165296

