---
title: "Miscellaneous Useful Functions"
output: html_document
---

Load libraries and data that will be used in this lesson,
```{r}
library("gapminder")
library("uwpols501")
data("gapminder")
```

# seq

The colon operator generates a sequence of integers b
```{r}
1:5
```

The function `seq` is a more flexible function to generate sequences,
```{r}
seq(1, 10, by = 10)
seq(1, 10, length.out = 20)
```
The function `seq_along` generates a sequence of integers the same length as the vector it is given,
```{r}
foo <- c("a", "b", "c")
seq_along(foo)
```
The function `seq_len` generates a sequence from 1 to the out,
```{r}
seq_len(10)
```

# which

The `which` function gives the indexes of a vector that fulfill a certain condition.
For example,
```{r}
which(gapminder$country == "United States")
which(gapminder$country == "United States" & gapminder$year > 2000)
which(gapminder$lifeExp < 32)
```

# match, %in%

The function `match` and `%in%` operator can be used to check which values in a vector 
are in another set of values.

The operator `x %in% y` returns `TRUE` if a value in `x` appears in `y`, and `FALSE` otherwise.
```{r}
letters %in% c("a", "b", "c")
gapminder$year %in% c(2007, 2002)
```
The function `match(x, table)` gives the index of the first match of `x` in `table` and `NA` otherwise.
```{r}
match(c("a", "b", "c", "alpha", "0"), letters)
match(gapminder$country, c("United States", "Mexico", "Canada"))
```


# cut

The function `cut` can be used to split a numeric vector into categories.
```{r, eval=FALSE}
cut(x, breaks)
```
creates a new vector with categories determined by the values of `breaks`.
If a single number is given for `breaks` it divides `x` into that number of categories:
```{r}
x <- 1:20
cut(x, 2)
cut(x, 5)
cut(x, 7)
```
If multiple values are given, `cut()` uses those values to split the vector into categories:
```{r}
cut(x, c(2, 10, 15))
```
Note that  `cut` splits the vector so that the values in each category does not include the lower value, but does include the upper value.
To reverse it so that the lower value is included, but the upper value is excluded, set the argument `right = TRUE`,
```{r}
cut(x, c(2, 10, 15), right = FALSE)
```
Also not that any elements in `x` less than the smallest break or greater than the largest break are set to `NA`.
To always create intervals that include all values smaller or lower, add `-Inf` and `Inf` to the breaks,
```{r}
cut(x, c(-Inf, 2, 10, 15, Inf))
```
You can add labels to the categories using the `labels` argument,
```{r}
cut(x, c(-Inf, 2, 10, 15, Inf), labels = c("Very Low", "Low", "High", "Very High"))
```


# mapvalues, revalue

The `mapvalues` and `revalue` functions allow you to easily replace values.
These functions are included in the `uwpols501` package (they are originally in the **plyr** package and re-exported for convenience because many functions in the **plyr** package have names that conflict with those in the **dplyr** package).

```{r}
x <- c("a", "b", "c")
mapvalues(x, c("a", "c"), c("A", "C"))

# Works on factors
y <- factor(c("a", "b", "c", "a"))
mapvalues(y, c("a", "c"), c("A", "C"))

# Works on numeric vectors
z <- c(1, 4, 5, 9)
mapvalues(z, from = c(1, 5, 9), to = c(10, 50, 90))

x <- c("a", "b", "c")
revalue(x, c(a = "A", c = "C"))
revalue(x, c("a" = "A", "c" = "C"))

y <- factor(c("a", "b", "c", "a"))
revalue(y, c(a = "A", c = "C"))
```

Using the gapminder data, replace the names of a few countries,
```{r}
country <- plyr::revalue(gapminder$country,
                                   c("United States" = "USA", 
                                     "United Kingdom" = "GBR",
                                     "Mexico" = "MEX",
                                     "Canada" = "CAN"))
unique(country)
```
or using `mapvalues`,
```{r}
country <- plyr::mapvalues(gapminder$country,
                         c("United States",
                           "United Kingdom",
                           "Mexico",
                           "Canada"),
                         c("USA", "GBR", "MEX", "CAN"))

unique(country)
```
