---
title: "Descriptive Statistics in R"
author: "Jeffrey Arnold"
date: "January 20, 2016"
output: html_document
---

```{r results='hide',echo=FALSE}
source("init.R")

```

This lesson provides examples of functions to calculate summary statistics in R.
It will use the gapminder. 

This lesson will also use the following packages,
```{r message=FALSE}
library("dplyr")
library("tidyr")
library("ggplot2")
```


# Gapminder Data

For this lesson, we will use the `gapminder` data in the **gapminder** package,
```{r}
library("gapminder")
data("gapminder")
```
For simplicity, we will focus on only the 2007 data,
```{r}
gapminder2007 <- gapminder[gapminder$year == 2007, ]
```
and save life expectancy and continent to their own vectors,
```{r}
lifeExp <- gapminder2007$lifeExp
continent <- gapminder2007$continent
```

# Miscellaneous Functions

R provides functions to calculate many common descriptive statistics:

- `mean(x)`: Mean 
- `median(x)`: median
- `sd(x)` : Standard deviation
- `var(x)` : Variance
- `IQR(x)` : Inter-quartile range
- `fivenumber(x)`: Tukey's five number summary
- `quantile(x, prob)` : Quantiles
- `range(x)`: Minimum and maximum
- `diff(range(x))` : Range

For life expectancy calculate the mean,
```{r}
mean(lifeExp)
```
median,
```{r}
median(lifeExp)
```
five number summary (minimum, lower-hinge, median, upper-hinge, maximum)
```{r}
fivenum(lifeExp)
```
standard deviation and variance,
```{r}
sd(lifeExp)
var(lifeExp)
```
inter-quartile range,
```{r}
IQR(lifeExp)
```
minimum and maximum,
```{r}
min(lifeExp)
max(lifeExp)
```
the min (0%), 1st quartile (25%), median (50%), 3rd quartile (75%), and maximum (100%),
```{r}
quantile(lifeExp)
```
arbitrary quantiles,
```{r}
quantile(lifeExp, prob = c(.33, .54, .86))
```

The `range` function returns the minimum and maximum, so take the difference of the range to calculate the range (maximum - minimum),
```{r}
range(lifeExp)
diff(range(lifeExp))
```

# Descriptive statistics by groups

The easiest way to calculate descriptive statistics by groups is to use **dplyr** `summarize()` with `group_by()`,
```{r}
gapminder %>%
  group_by(year) %>%
  summarize(lifeExp_mean = mean(lifeExp),
            lifeExp_median = median(lifeExp),
            gdpPercap_mean = mean(gdpPercap),
            gdpPercap_sd = sd(gdpPercap))
```
Alternatively, grouping by continent and year,
```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarize(lifeExp_mean = mean(lifeExp),
            lifeExp_median = median(lifeExp),
            gdpPercap_mean = mean(gdpPercap),
            gdpPercap_sd = sd(gdpPercap))
```

To calculate several descriptive statistics for several variables, the function
`summarize_each` is useful,
```{r}
gapminder %>%
  group_by(year) %>%
  summarize_each(funs(mean, median), lifeExp, gdpPercap)
```

Alternatively, if you want more control, you can combine `summarize` with `gather` and `spread`,
```{r}
gapminder_by_year <-
  gapminder %>%
    gather(variable, value, -year, -country, -continent) %>%
    group_by(year, variable) %>%
    summarize(mean = mean(value),
              sd = sd(value),
              median = median(value),
              max = max(value),
              min = min(value))
gapminder_by_year
```
If you need to convert back to a dataset where observations are years,
```{r}
gapminder_by_year %>%
  gather(stat, value, -year, -variable) %>%
  unite(varname, variable, stat) %>%
  spread(varname, value) %>%
  glimpse()
```


# Table and Mode

The `table` function counts the number of observations equal to a value,
```{r}
table(continent)
```
This can be used to find the `mode` for a discrete variable [^mode],
```{r}
continent_counts <- table(continent)
names(continent_counts)[continent_counts == max(continent_counts)]
```

- `table(continent)` counts the number of observations taking each value
- `names(continent_counts)` In the vector returned by the table command, the values of the observations are the names, and counts are values in the vector.
- `continent_counts == max(continent_counts)` is `TRUE` for all (possibly more than one) elements which have the largest counts

That code only works well for discrete data. 
The basic method used to find a mode of continuous is to find the maximum value of the kernel density function.
See package **modeest** function `mlv` for various methods to calculate the mode, including for methods to estimate the mode of continuous distributions.
```{r}
library("modeest")
mlv(lifeExp, method = "shorth")
```

[^mode]: R does not have a built-in function to calculate the mode of a sample distribution.
         The function `mode` in R is not related to the statistics, but is a programming function related to how the object is stored in memory.

# Skewness, Kurtosis

Generally you will not be interested in the higher moments of sample data, such as the skewness and kurtosis. If you are, the package **moments** provides functions for these.
```{r}
library("moments")
skewness(lifeExp)
kurtosis(lifeExp)
```



# Plots

Several plots are useful for descriptive statistics: boxplots, bar plots, histograms, density plots, and quantile plots.

## Bar-plot

A bar plot describes the distribution of a discrete-variable,
```{r}
ggplot(gapminder2007, aes(x = continent)) +
  geom_bar()
```
This could be faceted in order to compare across different categories,
```{r}
ggplot(gapminder, aes(x = cut(lifeExp, c(0, 40, 50, 60, 70, 80, Inf)))) +
  geom_bar() +
  facet_wrap(~ year, ncol = 1) +
  xlab("Life Expectancy categories")
```

## Histograms, Density

A histogram shows the distribution

```{r message = FALSE}
ggplot(gapminder2007, aes(x = lifeExp)) +
  geom_histogram()
```
By default, `geom_histogram` gives the message, if you do not choose the widths of the bins,
```
`stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
```
You can manually choose a binwidth to remove that message. 
In this case, 5 years sounds reasonable,
```{r}
ggplot(gapminder2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5)
```

We can compare across groups using faceting.
To compare the distributions across continents,
```{r}
ggplot(gapminder2007, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(~continent, ncol = 1)
```
or continents and years,
```{r}
ggplot(gapminder, aes(x = lifeExp)) +
  geom_histogram(binwidth = 5) +
  facet_grid(year ~ continent)
```
Another way to compare groups is to use color and transparency to overlap histograms, although this will not work for black and white publications,
```{r}
ggplot(gapminder2007, aes(x = lifeExp, fill = continent)) +
  geom_histogram(binwidth = 5, alpha = 0.3)
```
```{r}
ggplot(gapminder, aes(x = lifeExp, fill = continent)) +
  geom_histogram(alpha = 0.3, binwidth = 5) +
  facet_wrap(~year, ncol = 1)
```

Density plots are a continuous alternative to histograms.
```{r}
ggplot(gapminder2007, aes(x = lifeExp)) +
  geom_density()
```
It is often useful to combine a density plot with a rug plot, which shows a tick mark
for every value in the data,
```{r}
ggplot(gapminder2007, aes(x = lifeExp)) +
  geom_density() + 
  geom_rug()
```

We can compare across groups using faceting.
To compare the distributions across continents,
```{r}
ggplot(gapminder2007, aes(x = lifeExp)) +
  geom_density() +
  geom_rug() +
  facet_wrap(~continent, ncol = 1)
```
or continents and years,
```{r}
ggplot(gapminder, aes(x = lifeExp)) +
  geom_density() +
  geom_rug() +
  facet_grid(year ~ continent)
```
Another way to compare groups is to use color, although this will not work well for black and white publications,
```{r}
ggplot(gapminder2007, aes(x = lifeExp, color  = continent)) +
  geom_density()
```
```{r}
ggplot(gapminder, aes(x = lifeExp, color = continent)) +
  geom_density() +
  facet_wrap(~year, ncol = 1)
```

To calculate histograms for multiple variables, it is easiest to use `gather()` to make the dataset long, with one row per (country, variable),
```{r message = FALSE}
gapminder2007_long <- gather(gapminder2007, variable, value,
                             -country, -year, -continent)
ggplot(gapminder2007_long, aes(x = value)) +
  geom_histogram() +
  facet_wrap(~variable, ncol = 1, scales = "free")
```


## Boxplot

Boxplots summarizes the distribution into five values and indicates outliers,

- median: middle line
- 1st quartile: bottom of box
- 3rd quartile: top of box
- lowest datum within 1.5 IQR of the lower quartile: end of lower whisker
- highest datum within 1.5 IQR of the upper quartile: end of upper whisker

Boxplots are best used for comparing the distributions of several groups.
```{r}
ggplot(gapminder2007, aes(x = continent, y = lifeExp)) +
  geom_boxplot()
```
Note that `geom_boxplot` requires an `x` aesthetic. 
If you need to make a boxplot for only one group (which may not be a good idea),
```{r}
ggplot(gapminder2007, aes(x = 1, y = lifeExp)) +
  geom_boxplot() +
  xlab("")
```

More comparisons can be made by faceting over other categories,
```{r}
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  facet_wrap(~year, ncol = 1)
```

A continuous alternative to a boxplot is a violin plot, which can be thought of as a cross between a boxplot and a density plot,
```{r}
ggplot(gapminder2007, aes(x = continent, y = lifeExp)) +
  geom_violin()
```
```{r}
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_violin() +
  facet_wrap(~year, ncol = 1)
```


## QQ Plot

Quantile plots compare the empirical cumulative distribution function of a sample to a theoretical distribution (usually the normal).
These are often used to check the normality of sample data.
```{r}
ggplot(gapminder2007, aes(sample = lifeExp)) +
  geom_qq()
```
This plots the qq-plots of several variables in the gapminder data,
```{r}
ggplot(gather(gapminder2007, variable, value, -year, -continent, -country),
       aes(sample = value)) +
  geom_qq() +
  facet_wrap(~variable, ncol = 1, scales = "free_y")
```
None of these distributions are particularly normal.
However, transforming `gdpPercap` and `pop` makes them closer to normal,
```{r}
ggplot(mutate(gapminder2007, 
              gdpPercap = log(gdpPercap),
              pop = log(pop)) %>%
              gather(variable, value,
                     -year, -continent, -country),
       aes(sample = value)) +
  geom_qq() +
  facet_wrap(~variable, ncol = 1, scales = "free_y")
```




