---
title: "Descriptive Statistics in R"
author: "Jeffrey Arnold"
date: "January 20, 2016"
output: html_document
---

```{r results='hide',echo=FALSE}
source("init.R")
```


R provides functions to calculate many common descriptive statistics:

- `mean(x)`: Mean 
- `median(x)`: median
- `sd(x)` : Standard deviation
- `var(x)` : Variance
- `IQR(x)` : Interquartile range
- `fivenumber(x)`: Tukey's five number summary
- `quantile(x, prob)` : Quantiles
- `range(x)`: Minimum and maximum
- `diff(range(x))` : Range
- `table(x)` Counts of values

`r challenge_start()`

Using the `gapminder` data in the **gapminder** package,
```{r}
library("gapminder")
data("gapminder")
```

`r challenge_end()`

`r solution_start()`



`r solution_end()`


# Mode

R does not have a built-in function to calculate the mode of a sample distribution.
The function `mode` in R is not related to the statistics, but is a programming function related to how the object is stored in memory.

To calculate the mode for discrete data, find the values with the highest count
using `table`:
```{r}
tbl <- table(x)
names(tbl)[tbl == max(tbl)]
```

- `table(x)` counts the number of observations taking each value
- `names(tbl)` In the vector returned by the table command, the values of the observations are the names, and counts are values in the vector.
- `tbl == max(tbl)` is `TRUE` for all (possibly more than one) elements which have the largest counts

That code only works well for discrete data. 
See package **modeest** function `mlv` for various methods to calculate the mode, including for methods to estimate the mode of continuous distributions.

# Skewness, Kurtosis

Generally you will not be interested in the higher moments of sample data, such as the skewness and kurtosis. If you are, the package **moments** provides functions for these.
