---
title: Difference in Means. T-Distribution and Hypothesis Testing
date: January , 2016
author: Andreu Casas
---

```{r init,results='hide',echo=FALSE}
source("init.R")
```

Read the data in:
```{r}
library(rio)
data <- import('data/ajr_2001.dta')
```

`r challenge_start()`
Take a look at the data:

  - Are there any missing values?
  - How many observations with country names (`shortnam`) are in the dataset?
  - How many unique countries are in the dataset?
  - Is there any country that shows up twice?
`r challenge_end()`

```{r, eval=FALSE}
library(dplyr)
head(data)
nrow(na.omit(data))
head(data$shortnam)
length(data$shortnam)
```
```{r, results='hide'}
data %>%
  filter(shortnam != "") %>%
  nrow()
clean_data <- data %>%
  filter(shortnam != "") %>%
  filter(shortnam != ".")
```

```{r, eval=FALSE}
head(sort(table(clean_data$shortnam), decreasing = T))
```

The first 120 rows of the original dataset do not have a value for the country-name variable (`shortnam`) and they also have missing values for most of the other variables. Not clear what they are: getting rid of them.

When importing STATA (`.dta`) dataset using the rio package, we have access to the variable description. We wrote a function in the `uwpols501` package to check at the variable description of STATA datasets loaded into R using `rio`. 

```{r, eval=FALSE}
library(devtools)
install_github(username = "jrnold", repo = "UW-POLS501/r-uwpols501")
library(uwpols501)
extract_var_info(clean_data)
```

The variable `extmort4` contains information about *potential settlers mortality* (See Note in Figure 1 of the paper for a description of the variable: *measured in terms of deaths per annum per 1,000...*)

`r challenge_start()`
Explore the `extmort4` and `logem4` variable:

  - How many missing values do they have?
  - Do they have the same missing values? Why or why not?
  - What is their mean, median, first, and third quantile?
  - What would be the best way to plot this variable in order to analyze its distribution?
`r challenge_end()`

```{r, eval=FALSE}
summary(clean_data$extmort4)
summary(clean_data$logem4)
```

`logem4` is the log of `extmort4`. They calculate the log of mortality-rate because they theorize that it has a log-linear relationship with the type of institutions (`avexpr`: average expropriation risk). However, some observations have a value for the `logem4` variable but not for the `extmort4` one, which does not make much sense (looks like an error). Create a new variable (e.g. `log_mort`) with the log of `extmort4`.

```{r}
clean_data$log_mort <- log(clean_data$extmort4)
```


```{r, message=FALSE}
library(ggplot2)
p <- ggplot(clean_data, aes(x= extmort4))
p + geom_histogram(bins = 30)
p + geom_density() 
ggplot(clean_data, aes(x = log_mort)) + geom_density()
```

Use plot to explore the relationship between the variables `logem4` and `avexpr`. 

```{r}
ggplot(clean_data, aes(x = logem4, y = avexpr)) + 
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

Another way to evaluate their theory that different **institutions** (`avexpr`: average expropiation risk) were set up in places where **colonization** was highly costly than in places that were easy to colonize (`logem4`: the log of their mortality rate measure) would be to compare the **mean** institution (average expropiation risk) in countries with high colonization costs and in countries with low colonization costs.

Create a new dummy variable indicating those countries that have a value of `logem4` in the top quantile (==1). 

```{r}
clean_data <- mutate(clean_data, high_mort = as.numeric(logem4 > quantile(clean_data$logem4, probs = .75, na.rm = TRUE)))
```

Plot the distribution of the `logem4` variable for these two groups side-by-side.

```{r}
ggplot(filter(clean_data, high_mort == 1), aes(x = logem4)) +
  geom_density(fill= "green4", alpha = 0.3) +
  geom_density(data = filter(clean_data, high_mort == 0),
               aes(x = logem4), fill = "blue4", alpha = 0.3)

```




