---
title: R Directories
date: January 22, 2016
author: Andreu Casas
---

```{r init,results='hide',echo=FALSE}
source("init.R")
```

## Learning Objectives

Understand R directories and how to use them

### What is a `directory`?

- A directory is a **folder** in your computer
- All programs and data in your computer is in some directory
- Directories are hierarchically organized
```
main_dir
│   README.md
│   file1.txt    
│
└───folder1
    │   file11.txt
    │   file12.txt
    │
    ├───subfolder1
    │   │   file111.txt
    │   │   file112.txt
```
- All directories and files have a `path` that allows us to access that directory or file
```
"main_dir/folder1/subfolder1/file111.txt"
```

### R and the working directory

When you start an R session, R gets executed in a particular directory in the computer. We call that directory the `working directory`. You can know your working directory using the command `getwd()`:

If you want to change your working directory, you can use the function `setwd()`; which takes the path of the new working directory as an argument:

```{r, eval=FALSE}
setwd("/main_dir/folder1")
```

### Reading files in R

When reading data into R (e.g. a CSV file), if you only give the file name, R is going to look for that file in the current working directory. If the file is not there but it's in a subdirectory of the working directory, you need to provide a partial path to the file indicating the subdirectory strucutre to get to that file. For example, imagine that R is being executed in the `main_dir` of the previous diagram and you want to read the **file112.txt**. You would provide the path from `main_dir` to **file112.txt**:

```{r, eval=FALSE}
data <- read.table(file = "/folder1/subfolder1/file112.txt")
```

If the file is neither in the working directory or in a subdirectory of the working directory, you need to provide the full path to that file. For example, imagine that in this case R is being executed in the `subfolder1` directory and you want to read the **file1.txt**. You would do the following:

```{r, eval=FALSE}
data <- read.table(file = "/main_dir/file1.txt")
```



