---
title: "Distributions in R"
author: "Jeffrey Arnold"
date: "January 28, 2016"
output: html_document
---

```{r}

```


# Sampling



# Seed

# Distributions

- `r*` draw samples from a distribution
- `q*` the quantile function of a distribution: for a given $p$, the $x$ such that $P(X < x) = p$.
- `p*` the cumulative probability function of a distribution: for a given $x$, $P(X < x)$.
- `d*` the probability density (or mass) function of a distribution: $P(x)$.

## Normal


```{r echo=FALSE,results='hide'}
opts_chunk$set(tidy=FALSE,fig.height=3,fig.width=4,dev='cairo_pdf',message=FALSE)
set.seed(1235)
normal_prob_area_plot <- function(lb=-Inf, ub=Inf, mean = 0, sd = 1, limits = c(mean - 3 * sd, mean + 3 * sd)) {
    x <- seq(limits[1], limits[2], length.out = 100)
    xmin <- max(lb, limits[1])
    xmax <- min(ub, limits[2])
    areax <- seq(xmin, xmax, length.out = 100)
    area <- data.frame(x = areax, ymin = 0, ymax = dnorm(areax, mean = mean, sd = sd))
    (ggplot()
     + geom_line(data.frame(x = x, y = dnorm(x, mean = mean, sd = sd)),
                 mapping = aes(x = x, y = y))
     + geom_ribbon(data = area, mapping = aes(x = x, ymin = ymin, ymax = ymax))
     + scale_x_continuous("x", limits = limits)
     + scale_y_continuous("p(x)"))
}
```


# Normal distribution

This lab will focus on two functions associated with the normal distribution.

- ``pnorm`` Find the probability that the random variable is less than (or greater than) a given value.
- ``qnorm`` For a given probability, find the associated value of the random variable. This is called the quantile.

There are two other R functions that are less critical for the purposes of answering homework questions but are useful for understanding the frequency definition of probability.

- ``rnorm`` Draw random values from a distribution of the normal distribution.
- ``dnorm`` Find the density of a normal distribution at a given value. This is what is used to draw density plots.

This lab will use ggplot for plotting so load that package now.

```{r }
library("ggplot2")
```


## Probability Function

The function pnorm returns $p(X \leq x)$ for a normal variable $X$ and a value $x$.
This function used to find the tail probabilities for the standard normal distribution ($z$-scores), or more generally, for the values of any normal distribution.

Consider a standard normal distribution ($X \sim N(0, 1)$).
To find $P(X \leq 1.96)$,

```{r }
pnorm(1.96)
```


Visually, ``pnorm(1.96)`` returns the value of the shaded area in the following:


```{r fig.cap="The area $P(X \\leq 1.96)$ is about 0.975",echo=FALSE}
normal_prob_area_plot(-Inf, 1.96)
```

\clearpage

To find $P(X > 1.96)$ you can either use rules of probability, $P(X > 1.96) = 1 - P(X \leq 1.96)$,

```{r }
1 - pnorm(1.96)
```

or set the option ``lower.tail`` to ``FALSE``,

```{r }
pnorm(1.96, lower.tail = FALSE)
```

\clearpage

Visually, ``pnorm(1.96, lower.tail=FALSE)`` returns the value of the shaded area in the following plot

```{r fig.cap="The area $P(X \\geq 1.96)$ is about 0.025",echo=FALSE}
normal_prob_area_plot(1.96)
```

For an arbitrary normal distribution (one in which the mean does not equal 0 or the standard deviation does not equal 1), you can either (1) convert the value to a z-score and then use ``pnorm`` as shown above, or (2) use ``pnorm`` with the ``mean`` and ``sd`` options.

Consider a normal distribution with mean 2 and standard deviation 4.
Find $P(X \leq 8)$.
The first method is to convert to a z-score and then use ``pnorm``.

```{r }
z <- (8 - 2) / 4
pnorm(z)
```

The second method is to use the ``mean`` and ``sd`` options in ``pnorm``

```{r }
pnorm(8, 2, 4)
```

As expected, they produce the same answer.


## Quantile Function

If you are given a probability and asked to find a value or a z-score, use ``qnorm``.

Consider a standard normal distribution ($X \sim N(0, 1)$).
Suppose you want to know the value at the 97.5 percentile.
Slightly more formally, you want to find the z-score such that $P(X \leq z) = 0.975$.

```{r }
qnorm(0.975)
```


Notice that ``qnorm`` produces the inverse (opposite) of ``pnorm``.
Given a value in the distribution, ``pnorm`` gives you a probability.
Given a probability, ``qnorm`` gives you a value.

One way of seeing this is by taking the ``qnorm`` of the value returned by ``pnorm``.
It will be exactly the probability that was first given as an argument to ``pnorm``.
For example, in the following code, ``qnorm(0.975)`` finds the z-score such that $P(X \leq z) = 0.975$ which is 1.959964.
Then ``pnorm`` finds the probability that $X$ is less than the $z$-score of 1.959964, which is 0.975.

```{r }
pnorm(qnorm(0.975))
```



To find a z-score such that a given fraction of the distribution is above it, you can either use rules of probability, $P(X > x) = 1 - P(X \leq x)$, or, as in ``pnorm``, set the option ``lower.tail=FALSE``.
For example, find the value such that 0.025 of the distribution is greater than it.
This can be done by,

```{r }
qnorm(1 - 0.025)
```

or

```{r }
qnorm(0.025, lower.tail = FALSE)
```


Now consider the random variable $Y$ distributed with mean 2 and standard deviation 4.
What is the value of $Y$ greater than 93.3 percent of the distribution?
This can be calculated in two ways.
First, by calculating the z-score and then transforming it.

```{r }
z <- qnorm(0.933)
2 + z * 4
```


Second, by using the ``mean`` and ``sd`` options in ``qnorm``,


```{r }
qnorm(0.933, 2, 4)
```


## Sampling and Density

You can use the ``rnorm`` function to draw a sample from a normal distribution.

```{r }
rnorm(5)
```

This draws a sample of size 5 from a normal distribution with mean 1 and standard deviation 2.

```{r }
rnorm(5, 1, 2)
```

The following code compares the kernel density of a sample drawn from the normal distribution with the density function of the normal distribution.

```{r }
library("ggplot2")
```


This draws as sample of size ``nsample`` with mean ``mu``, with standard deviation ``sigma``.

```{r }
mu <- 1 ## mean
sigma <- 2 ## standard deviation
nsample <- 50 ## sample size
smpl <- rnorm(nsample, mu, sigma)
```


This code will plot the kernel density of the normal distribution sample with the density function of the normal distribution.
You don't need to worry about exactly how this code works.
The lines below the density plot are called a "rug" and show the location of the values used in the density plot.


```{r normal_density,fig.cap = "Kernel Density of a Sample from the Normal Distribution and the Density Function of the Normal Distribution"}
x <- seq(mu - 3 * sigma, mu + 3 * sigma, length.out=100)
y <- dnorm(x, mu, sigma)
(qplot(smpl, geom = c("density", "rug"), sides = "b",
       xlab = "x")
 + geom_line(data = data.frame(x = x, y = y),
             mapping = aes(x = x, y = y),
             color = "red"))
```


How does the kernel density plot of the sample compare with the density function of the normal distribution as the sample size increases?
Change the value of ``nsample`` and rerun the code.
Try values of 10, 50, 100, 500, 1000, 10000.
You will see that the kernel density of the sample approaches the density function of the normal distribution as the sample size gets large.
This is what is meant by the frequency definition of probability in Agresti and Finlay.



## Bernoulli and Binomial

`dbinom` calculates the density of a binomial distribution.
```{r}
.data <- 
  data_frame(x = 0:10,
             y = dbinom(x, size = max(x), prob = 0.3))
ggplot(.data, aes(x = x, y = y)) +
  geom_bar(stat = "identity")
```

```{r}
p <- 0.3
size <- 10
pmf <- 
  data_frame(x = 0:size,
             y = dbinom(x, size = size, prob = p))
smpl <-
  data_frame(x = rbinom(100, size = size, prob = p)) %>%
  group_by(x) %>%
  tally() %>%
  mutate(freq = n / sum(n))
ggplot() +
  geom_bar(data = pmf, mapping = aes(x = x, y = y),
           stat = "identity", alpha = 0.3, fill = "red") +
  geom_bar(data = smpl, mapping = aes(x = x, y = freq),
           stat = "identity", alpha = 0.3)
```

- What happens as `prob` changes?
- What happens as `size` changes?


## Geometric

Function `dgeom`

```{r}
p <- 0.3
smpl <-
  data_frame(x = rgeom(100, prob = p)) %>%
  group_by(x) %>%
  tally() %>%
  mutate(freq = n / sum(n))
pmf <- 
  data_frame(x = 0:max(smpl$x),
             y = dgeom(x, prob = p))
ggplot() +
  geom_bar(data = pmf, mapping = aes(x = x, y = y),
           stat = "identity", alpha = 0.3, fill = "red") +
  geom_bar(data = smpl, mapping = aes(x = x, y = freq),
           stat = "identity", alpha = 0.3) +
  xlim(0, qgeom(0.995, prob = 0.3))
```

## Negative Binomial

Function `dnbinom`

```{r}
p <- 0.3
size <- 5
xmax <- qnbinom(0.995, size = size, prob = p)
smpl <-
  data_frame(x = rnbinom(1000000, size = size, prob = p)) %>%
  group_by(x) %>%
  tally() %>%
  mutate(freq = n / sum(n))
pmf <- 
  data_frame(x = 0:xmax,
             y = dnbinom(x, size = size, prob = p))
ggplot() +
  geom_bar(data = pmf, mapping = aes(x = x, y = y),
           stat = "identity", alpha = 0.3, fill = "red") +
  geom_bar(data = smpl, mapping = aes(x = x, y = freq),
           stat = "identity", alpha = 0.3) +
  xlim(0, xmax)
```

- what happens when change size? when change 
- if size = 1, it is the geometric distribution

## Poisson

Function `dpois`

```{r}
lambda <- 3
xmax <- qpois(0.995, lambda = lambda)
smpl <-
  data_frame(x = rpois(100, lambda = lambda)) %>%
  group_by(x) %>%
  tally() %>%
  mutate(freq = n / sum(n))
pmf <- 
  data_frame(x = 0:xmax,
             y = dpois(x, lambda = lambda))
ggplot() +
  geom_bar(data = pmf, mapping = aes(x = x, y = y),
           stat = "identity", alpha = 0.3, fill = "red") +
  geom_bar(data = smpl, mapping = aes(x = x, y = freq),
           stat = "identity", alpha = 0.3) +
  xlim(0, xmax)
```

- What happens as `lambda` increases?
- Compare `rbinom` with small `prob` and large `size` to a Poisson with `lambda = prob * size`.


# Quantile-Quantile Plot

In `ggplot2`, compare the quantiles of a sample to the quantile function of a distribution (often the normal distribution).

```{r}
ggplot(mtcars) +
  geom_qq(aes(sample = mpg)) +
  geom_abline(slope = 1, intercept = 0)
```

```{r}
ggplot(mtcars) +
  stat_qq(aes(sample = mpg, colour = factor(cyl)))

```


