---
title: "Random Sampling and Distributions in R"
author: "Jeffrey Arnold"
date: "January 28, 2016"
---

```{r results='hide', echo=FALSE, message = FALSE}
source("init.R")
set.seed(12350789)

knitr::opts_chunk$set(cache = FALSE)
```

This lesson covers sampling and distributions in R. Although all the functions
for the distributions are included in base R, several packages are used for
plotting and data wrangling,
```{r message = FALSE}
library("ggplot2")
library("tidyr")
library("dplyr")
```


# Simple Random Samples

The `sample` function draws a simple random sample.
```{r eval=FALSE}
sample(x, size, replace = FALSE, prob = NULL)
```
The function `sample` draws a sample of size `size` from the set in `x`.
If `replace = TRUE`, then it samples with replacement, otherwise, it samples without replacement.
The samples are drawn with equal probability, unless values are provided to `prob`.

To draw a sample of size 5 from the integers 1 to 10, without replacement.
```{r}
sample(1:10, 5)
```
If no size is given, it draws a sample of the same size as `x`.
Since by default, it samples without replacement, this is equivalent to randomly reordering the vector.
```{r}
sample(1:10)
```
To sample with replacement, add the argument `replace = TRUE`,
```{r}
sample(1:10, 5)
```
If `replace = TRUE`, you can draw samples with sizes larger than the vector of choices,
```{r}
sample(1:10, 20, replace = TRUE)
```

By default, `sample()` samples from the elements in `x` with equal probability.
To sample from the elements in `x` with unequal probability, provide those probabilities with the `prob` argument.
For example, to sample from a Bernoulli distribution where the probability of 1 is 0.7 and the probability of 0 is 0.3,
```{r}
sample(1:0, 10, replace = TRUE, prob = c(0.7, 0.3))
```
Or to sample from a discrete distribution where $P(X = 1) = 0.2$, $P(X = 2) = 0.4$,
$P(X = 3) = 0.3$, $P(X = 4) = 0.1$,
```{r}
sample(1:4, 20, replace = TRUE, prob = c(0.2, 0.4, 0.3, 0.1))
```

Plotting the frequencies of samples without and with the `prob` argument shows
how the weighting affects the sampling,
```{r}
x_sample_unwt <-
  data_frame(x = sample(1:4, 10000, replace = TRUE)) %>%
  group_by(x) %>%
  tally() %>%
  mutate(prob = n / sum(n))

ggplot(x_sample_unwt, aes(x = x, y = prob)) +
  geom_bar(stat = "identity")
```
```{r}
x_sample_wt <-
  data_frame(x = sample(1:4, 10000, replace = TRUE,
                        prob = c(0.2, 0.4, 0.3, 0.1))) %>%
  group_by(x) %>%
  tally() %>%
  mutate(prob = n / sum(n))

ggplot(x_sample_wt, aes(x = x, y = prob)) +
  geom_bar(stat = "identity")
```



# Distributions

R includes

- `r*` draw samples from a distribution
- `q*` the quantile function of a distribution: for a given $p$, the $x$ such that $P(X < x) = p$.
- `p*` the cumulative probability function of a distribution: for a given $x$, $P(X < x)$.
- `d*` the probability density (or mass) function of a distribution: $P(x)$.

For example, the functions for the normal distribution are `rnorm`, `qnorm`, `pnorm`, and `dnorm`.

For the full list of distributions included in R see the help page for [Distributions](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Distributions.html).
For even more distributions, see the CRAN task view [Distributions](https://cran.r-project.org/web/views/Distributions.html).

## Normal Distribution

The functions for the normal distributions are

- `dnorm`: density function
- `pnorm`: probability function
- `qnorm`: quantile function
- `rnorm`: sampling function

The function `dnorm` calculates the density of the normal distribution at a point.
This can be used to draw the "bell-shaped" normal distribution curves.
For example,
```{r}
norm_density <- data_frame(x = seq(-3, 3, length.out = 100),
                           y = dnorm(x))
ggplot(norm_density, aes(x = x, y = y)) +
  geom_line()
```
By default, `dnorm` calculates the density for a standard normal distribution, $\mu = 0$ and $\sigma = 1$.
Use the arguments, `mean` and `sd`, to specify different means or standard deviations.
For example, to calculate the density function of a normal distribution with mean 2 and standard deviation 3,
```{r}
norm_density <- data_frame(x = seq(-7, 11, length.out = 100),
                           y = dnorm(x, mean = 2, sd = 3))
ggplot(norm_density, aes(x = x, y = y)) +
  geom_line()
```

This draws the PDF of several normal distributions with $\mu = 0$, and $\sigma \in \{1/3, 1/2, 0, 2, 3\}$
```{r}
expand.grid(mu = 0,
            sigma = c(1/2, 1, 2)) %>%
  group_by(mu, sigma) %>%
  do(data_frame(x = seq(-6, 6, length.out = 500),
                density = dnorm(x, mean = .$mu, sd = .$sigma))) %>%
  ggplot(aes(x = x, y = density, color = factor(sigma))) +
  geom_line()
```
This draws the PDF of several normal distributions with different means, $\mu \in \{-1, 0, 1\}$, and the same standard deviation, $\sigma = 1$,
```{r}
expand.grid(mu = c(-1, 0, 1),
            sigma = 1) %>%
  group_by(mu, sigma) %>%
  do(data_frame(x = seq(-6, 6, length.out = 500),
                density = dnorm(x, mean = .$mu, sd = .$sigma))) %>%
  ggplot(aes(x = x, y = density, color = factor(mu))) +
  geom_line()
```

The function `pnorm` calculates the cumulative distribution function for the normal distribution, $F(x) = \Pr(X \leq x)$.
For example, the probability that a normally distributed $X$ is less than -1.96 is about 0.025.
```{r}
pnorm(-1.96)
```
It can be given several values,
```{r}
round(pnorm(-3:3), 3)
```
Use the arguments `mean` and `sd` to set the the mean and standard deviation of the normal distribution.
E.g. the CDF of a normal distribution with $\mu = 1$ and $\sd = 2$,
```{r}
x <- -5:7
pnorm(seq(-5, 7, by = 2), mean = 1, sd = 2)
```
This is equivalent to using `pnorm` on the z-scores,
```{r}
mu <- 1
sigma <- 2
x <- -5:7
z <- (x - mu) / sigma
pnorm(z)
```
To calculate the values of the complementary cumulative distribution function, $\bar{F}(x) = \Pr(X > x)$, use the argument `lower.tail = FALSE` with `pnorm()`.
For example, the probability that a standard normal random variable $X$ is greater than -1.96 is,
```{r}
pnorm(-1.96, lower.tail = FALSE)
```
This is equivalent to
```{r}
1 - pnorm(-1.96)
```
You can calculate areas using `pnorm` as well.
For example, to calculate the probability $\Pr(-1.96 \leq X \leq 1.96)$,
```{r}
pnorm(1.96) - pnorm(-1.96)
```
As expected, this is approximately 95%.

The quantile function of the normal distribution gives for a given probability $p$, the value $x$ of the distribution such that the the probability that $\Pr(X \leq x) = p$.
The function `qnorm` calculates the normal distribution quantile function.
For example, the value of the normal distribution such that $\Pr(X < x) = .975$,
```{r}
qnorm(0.975)
```
As expected, this is approximately 97.5%.
```{r}
round(qnorm(c(0.0015, 0.025, 0.16, 0.5, 0.84, 0.975, 0.9985)), 2)
```

To sample from a normal distribution, use `rnorm()`,
```{r}
smpl <- rnorm(500)
ggplot(data_frame(x = smpl), aes(x = x)) +
  geom_density() +
  geom_rug() +
  stat_function(color = "red", fun = "dnorm")
```
Like the other normal distribution functions, it has `sd` and `mean` arguments,
```{r}
mu <- 2
sigma <- 4
smpl <- rnorm(500, mean = mu, sd = sigma)
ggplot(data_frame(x = smpl), aes(x = x)) +
  geom_density() +
  geom_rug() +
  stat_function(color = "red", fun = "dnorm", args = list(sd = sigma,
                                                          mean = mu))
```

## Student's t Distribution

- `r*` draw samples from a distribution
- `q*` the quantile function of a distribution: for a given $p$, the $x$ such that $P(X < x) = p$.
- `p*` the cumulative probability function of a distribution: for a given $x$, $P(X < x)$.
- `d*` the probability density (or mass) function of a distribution: $P(x)$.

For example, the functions for the normal distribution are `rnorm`, `qnorm`, `pnorm`, and `dnorm`.

For the full list of distributions included in R see the help page for [Distributions](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Distributions.html).
For even more distributions, see the CRAN task view [Distributions](https://cran.r-project.org/web/views/Distributions.html).

## Normal Distribution

The functions for the normal distributions are

- `dt`: density function
- `pt`: probability function
- `qt`: quantile function
- `rt`: sampling function

The function `dt` calculates the density of the Student's t distribution.
For example, the density of the Student's t distribution with 10 degrees of freedom is,
```{r}
data_frame(x = seq(-3, 3, length.out = 100),
                           y = dt(x, 10)) %>%
ggplot(aes(x = x, y = y)) +
  geom_line()
```

This draws the PDF of several Student's t distributions with $df = \{1, 2, 4, 8, 16, 32, 64, Inf\}$
```{r}
data_frame(df = c(1, 2, 4, 8, 16, 32, 64, 512)) %>%
  group_by(df) %>%
  do(data_frame(x = seq(-6, 6, length.out = 500),
                density = dt(x, .$df))) %>%
  ggplot(aes(x = x, y = density, color = factor(df))) +
  geom_line()
```
Setting `df = Inf` (infinity) in a Student's t distribution gives the Normal distribution.

The function `pt` calculates the cumulative distribution function for the Student's t distribution, $F(x) = \Pr(X \leq x)$.
For example, the probability that a Student's t random variable $X$ is less than -1.96 will be somewhat higher than the 0.025 probability in a Normal distribution, and decrease with the degrees of freedom,
```{r}
pt(-1.96, df = c(1, 8, 16, 32, 512, Inf))
pnorm(-1.96)
```
To calculate the values of the complementary cumulative distribution function, $\bar{F}(x) = \Pr(X > x)$, use the argument `lower.tail = FALSE` with `pt()`.
For example, the probability that a Student's t distributed random variable $X$ is greater than -1.96 is,
```{r}
pt(-1.96, df = c(1, 8, 16, 32, 512, Inf), lower.tail = TRUE)
pnorm(-1.96)
```
This is equivalent to
```{r}
1 - pt(-1.96, df = c(1, 8, 16, 32, 512, Inf))
```
You can calculate areas using `pt` as well.
For example, to calculate the probability $\Pr(-1.96 \leq X \leq 1.96)$,
```{r}
degrees_freedom <- c(1, 8, 16, 32, 512, Inf)
pt(1.96, df = degrees_freedom) - pt(-1.96, df = degrees_freedom)
```
As expected, this is approximately 95%.

The quantile function of the Student's t distribution gives for a given probability $p$, the value $x$ of the distribution such that the the probability that $\Pr(X \leq x) = p$.
The function `qt` calculates the Student's t distribution quantile function.
For example, the value of the Student's t distribution such that $\Pr(X < x) = .975$, for several degrees of freedom, is
```{r}
qt(0.975, df = c(1, 8, 16, 32, 512, Inf))
```
As the degrees of freedom get larger, the .975 quantile approaches the normal distribution value of approximately 1.96.

To sample from a normal distribution, use `rt()`.
This samples from a Student's t distribution with 5 degrees of freedom, and compares the samples to the Student's t density function, and the normal distribution density function,
```{r}
degrees_freedom <- 5
smpl <- rt(1000, df = degrees_freedom)
ggplot(data_frame(x = smpl), aes(x = x)) +
  geom_density() +
  geom_rug() +
  stat_function(color = "blue", fun = "dt", args = list(df = degrees_freedom)) +
  stat_function(color = "orange", fun = "dnorm")
```


# Bernoulli, Binomial

The binomial distribution is a discrete distribution with the density function,
$$
p(x | n, p) = \choose{n}{x} p^x (1 - p)^{n - x}
$$
for $x \in 1, \dots, n$, and where $n$ is an integer $1, \dots$, and $p \in [0, 1$.
A Bernoulli distribution is a binomial distribution where $n = 1$,
$$
p(x | p) = p^x (1 - p)^{1 - x}
$$
for $x \in \{0, 1\}$.

The functions for the Binomial distributions are

- `dbinom`: density function
- `pbinom`: probability function
- `qbinom`: quantile function
- `rbinom`: sampling function

To calculate the Binomial probability density function use `dbinom`.
For example, this calculates the density for a binomial distribution with $p = 0.5$
and $n = 10$,
```{r}
dbinom(0:10, size = 10, prob = 0.5)
```

This plots the Binomial probability density function for $p = 0.5$ and $n \in \{5, 10, 15, 20\}$,
```{r}
binom_densities <-
  expand.grid(size = c(5, 10, 15, 20),
              prob = 0.5) %>%
  group_by(size, prob) %>%
  do(data_frame(x = 0:20,
                density = dbinom(x, size = .$size, prob = .$prob)))

ggplot(binom_densities, aes(x = x, y = density, colour = factor(size))) +
  geom_point() +
  geom_line()
```

This plots the Binomial probability density function for $p \in \{0.125, 0.25, 0.5, 0.75\}$ and $n = 20$,
```{r}
binom_densities <-
  expand.grid(size = 20,
              prob = c(0.125, 0.25, 0.5, 0.75)) %>%
  group_by(size, prob) %>%
  do(data_frame(x = 0:20,
                density = dbinom(x, size = .$size, prob = .$prob)))

ggplot(binom_densities, aes(x = x, y = density, colour = factor(prob))) +
  geom_point() +
  geom_line()
```

The function `pnorm()` calculates the Binomial cumulative probability function, $F(x) = \Pr(X \leq x)$,
```{r}
pbinom(0:5, size = 5, prob = 0.5)
```
which is equivalent to cumulatively summing the density function,
```{r}
cumsum(dbinom(0:5, size = 5, prob = 0.5))
```
To calculate the complementary cumulative probability function (CCDF), $\bar{F}(x) = \Pr(X > x)$,
use the argument `lower.tail = FALSE`,
```{r}
pbinom(0:5, size = 5, prob = 0.5, lower.tail = FALSE)
```
This is equivalent to 1 minus the CDF,
```{r}
1 - pbinom(0:5, size = 5, prob = 0.5)
```


This plots the Binomial cumulative distribution function for $p = 0.5$,
and $n \in \{5, 10, 15, 20\}$,
```{r}
expand.grid(size = c(5, 10, 15, 20),
            prob = 0.5) %>%
  group_by(size, prob) %>%
  do(data_frame(x = 0:20,
                density = pbinom(x, size = .$size, prob = .$prob))) %>%
  ggplot(aes(x = x, y = density, colour = factor(size))) +
    geom_point() +
    geom_line()
```

This plots the Binomial cumulative distribution function for $p = 0.5$, $p \in \{0.125, 0.25, 0.5, 0.75\}$ and $n = 20$,
```{r}
expand.grid(size = 20,
            prob = c(0.125, 0.25, 0.5, 0.75)) %>%
  group_by(size, prob) %>%
  do(data_frame(x = 0:20,
                density = pbinom(x, size = .$size, prob = .$prob))) %>%
  ggplot(aes(x = x, y = density, colour = factor(prob))) +
  geom_point() +
  geom_line()
```

The quantile function is the inverse of the cumulative density function. For a given probability, $p$, it returns the value $x$ such that $P(X \leq x) = p$.
For example, for the Binomial distribution with $n = 5$, and $p = 0.25$, the value where
$P(X \leq x) = 0.75$ is
```{r}
qbinom(0.75, size = 5, prob = 0.25)
```

This plots the Binomial quantile function for $p = 0.5$,
and $n \in \{5, 10, 15, 20\}$,
```{r}
expand.grid(size = c(5, 10, 15, 20),
            prob = 0.5) %>%
  group_by(size, prob) %>%
  do(data_frame(x = seq(0, 1, length.out = 101),
                density = qbinom(x, size = .$size, prob = .$prob))) %>%
  ggplot(aes(x = x, y = density, colour = factor(size))) +
    geom_point() +
    geom_line()
```

This plots the Binomial quantile function for $p = 0.5$, $p \in \{0.125, 0.25, 0.5, 0.75\}$ and $n = 20$,
```{r}
expand.grid(size = 20,
            prob = c(0.125, 0.25, 0.5, 0.75)) %>%
  group_by(size, prob) %>%
  do(data_frame(x = seq(0, 1, length.out = 101),
                density = qbinom(x, size = .$size, prob = .$prob))) %>%
  ggplot(aes(x = x, y = density, colour = factor(prob))) +
  geom_point() +
  geom_line()
```


## Geometric

TODO

Function is `dgeom`.

## Poisson

TODO

Function is `dpois`.


# Seed

When R generates random numbers it is not actually, random, but a type of algorithm
known as a pseudo-random number generator. In short, these are algorithms that
are deterministic, but produce a sequence of numbers that have properties similar to
those if they were actually generated randomly. This also means that you can
exactly replicate random number results in R, as long as you set an object called
the "seed" using a function called `set.seed`.

```{r}
set.seed(12345)
sample(1:10, 3)
sample(1:10, 3)
set.seed(12345)
sample(1:10, 3)
sample(1:10, 3)
set.seed(12345)
sample(1:10, 3)
sample(1:10, 3)
```

In scripts where you randomly sample (or to be safe, all scripts), you should set
a seed so that others can exactly replicate your results.
