---
title: "Exporting tables from R"
author: "Andreu Casas"
date: "March 4, 2016"
---

```{r init,results='hide',echo=FALSE}
source("init.R")
```

We often want to add in our papers tables that we have produced using R. That is an easy task if we are using R-Markdown, because we can simply add a chunk of code with that table to the Markdown report. However, how can we export an R table so we can add it to a Word document? In this lesson we'll see how to do so using three different packages:

- `htmlTables()`
- `xtable()`

### `htmlTable()`

We need an exmaple table we can use to practice. Let's fit a linear model using the dataset `iris` and then create a table with the output of the regression. We'll use the `tidy()` function of the `broom()` package to extract the coefficients table from the regression output.
 
```{r}
library(broom)
reg <- lm(Sepal.Length ~ Sepal.Width, data = iris )
reg_table <- tidy(reg)
reg_table
```

The values in the table have numerous decimals. Let's round them so we can fit them in a table. Notice that we can't round the first column because it's a character variable with the variable names.

```{r}
reg_table[-1] <- round(reg_table[-1],2)
reg_table
```

We can now use the `htmlTable()` function from the `htmlTable` package to print the table in the R-Studio viewer. Then we will be able to copy it and paste it in a Word document. In this case I'm selecting not to include the row names in the table (`rnames = FALSE`).

```{r, eval = FALSE}
htmlTable(reg_table, rnames = FALSE)
```

---

### `xtable()`

We can also use the `xtable()` package to create an html file that then we can open in a browser, copy the table we find in there, and paste it in a word document.

```{r, eval = FALSE}
library(xtable)
print(xtable(reg_table), type = "html", file = "reg_table_xtable.html")
```


