---
title: "Dplyr and tidyr tutorial with gapminder"
author: "Jeffrey Arnold"
date: "January 26, 2016"
output: html_document
---

```{r message=FALSE}
library("gapminder")
library("dplyr")
library("tidyr")
library("ggplot2")
```

```{r}
data("gapminder")
```

```{r echo=FALSE,results='hide'}
knitr::opts_chunk$set(results="hide", error=TRUE)
```


# `data_frame`

The `gapminder` data is special type of data frame created by the **dplyr** package,
```{r}
class(gapminder)
```
You can tell this because it has class "tbl_df" in addition to "data.frame".
`tbl_df` objects are mostly compatible with `data.frame` objects; 
you can use `[`, `[[`, `$`, and all the functions compatible with data frames.
The most noticeable way that they differ from `data.frame` objects is that they have prettier printing.
Instead of printing the whole object, it will only print the first few lines and columns,
```{r}
gapminder
```
Compare with how this prints if it is converted to a `data.frame`,
```{r results='hide'}
as.data.frame(gapminder)
```

`tbl_df` objects can be created from an existing data frame using the `tbl_df` function, or from scratch with the `data_frame` function.

You can read more about data frames in the **dplyr** package [here](https://cran.r-project.org/web/packages/dplyr/vignettes/data_frames.html).

# Glimpse

The common method to view an object in R is the `str` function
```{r}
str(gapminder)
```
The `glimpse` function in **dplyr** is similar, but specialized to data frames,
```{r}
glimpse(gapminder)
```


# Select rows with `filter()` or `slice()`

The functions `filter` and `slice` select a subset of rows in a data frame.

Select all rows in 2007,
```{r results='hide'}
filter(gapminder, year == 2007)
```
Select all rows from the US
```{r results='hide'}
filter(gapminder, country == "United States")
```
Select all observations in 2002 or 2007,
```{r}
filter(gapminder, year %in% c(2002, 2007))
```

Multiple arguments, are equivalent to "and".
Select all rows with `lifeExp` less than 50 and population more than 50 million, and in Asia,
```{r}
filter(gapminder, lifeExp < 50, pop > 50e+6, continent == "Asia")
```
The previous code is equivalent to,
```{r}
filter(gapminder, lifeExp < 50 & pop > 50e+6 & continent == "Asia")
```
Select all rows with an average life expectancy less than 50 or GDP per capita less than 365, in Asia, for years after 1970,
```{r}
filter(gapminder, lifeExp < 50 | gdpPercap < 456, continent == "Asia", year > 1970)
```

Select the first five rows,
```{r}
slice(gapminder, 1:5)
```
Select the the last row, using the special function `n()` within `slice`,
```{r}
slice(gapminder, n())
```
Select the rows 6 to 10,
```{r}
slice(gapminder, 6:10)
```

# Reorder rows with `rename()`

Arrange sorts rows by a variable.
Sort by life expectancy from lowest to highest,
```{r}
arrange(gapminder, lifeExp)
```
Sort by life expectancy from highest to lowest (descending values),
```{r}
arrange(gapminder, desc(lifeExp))
```

You can sort by multiple columns.
Sort by year (ascending), and life expectancy (descending),
```{r}
arrange(gapminder, year, desc(lifeExp))
```

# Select columns with `select()`

The `select` function selects columns.

Select columns by name
```{r}
select(gapminder, country, year, lifeExp)
```
Select all columns between `country` and `pop`,
```{r}
select(gapminder, country:pop)
```
Select all columns *except* continent,
```{r}
select(gapminder, -continent)
```
Reorder columns so that year is first, country is second, with the remaining columns in the same order,
```{r}
select(gapminder, year, country, everything())
```

# Extract distinct (unique) rows with `distinct()`

The function `distinct` returns the unique row values in a table.

This selects all unique continent, year values
```{r distinct1}
distinct(select(gapminder, continent, year))
```
This selects all distinct countries,
```{r}
distinct(select(gapminder, country))
```

# Add new columns with `mutate()`

Add the columns GDP, log GDP per capita, log population, and log GDP,
and a dummy variable for North America, 
```{r mutate1}
mutate(gapminder,
       GDP = pop * gdpPercap,
       log_pop = log(pop),
       log_gdpPercap = log(gdpPercap),
       log_gdp = log_pop + log_gdpPercap,
       na_eu = continent %in% c("United States", "Mexico", "Canada"))
```

# Rename columns with `rename()`

Rename columns with the `rename()` variable.
Rename `pop` to `population`, `gdpPercap` to `gdp_per_cap`, and `lifeExp` to `life_expectancy`,
```{r}
rename(gapminder,
       population = pop,
       gdp_per_cap = gdpPercap,
       life_expectancy = lifeExp)
```


# Randomly sample rows with `sample_n()` and `sample_frac()`

Use `sample_n` and `sample_frac` to sample rows.
Sample 10 rows,
```{r sample1}
sample_n(gapminder, 10)
```
Sample 5 percent of rows,
```{r}
sample_frac(gapminder, 0.05)
```
Sample all rows with replacement (bootstrap),
```{r}
sample_frac(gapminder, replace = TRUE)
```

# Summarize values with `summarise()`

The function `summarise` (or `summarize`) collapses a data frame to a single row, summarized by functions.

Create a data frame with the minimum and maximum years, the minimum, maximum, and median life expectancy, the average population, and the average GDP per capita, total number of observations, number of unique years, and number of unique countries,
```{r}
summarise(gapminder,
          year_min = min(year),
          year_max = max(year),
          life_exp_min = min(lifeExp),
          life_exp_max = max(lifeExp),
          life_exp_median = median(lifeExp),
          pop = median(pop),
          gdpPercap = mean(gdpPercap),
          nobs = length(country),
          n_years = length(unique(year)),
          n_country = length(unique(continent)))
```

# Grouped Functions

Use the `group_by` function to group the data frame.
This allows for calculating variables and 

For each country, calculate its average, median, and maximum GDP per capita, and median life expectancy,
```{r grouped1}
gapminder %>%
  group_by(country) %>%
  summarise(gdpPercap_mean = mean(gdpPercap),
            gdpPercap_median = median(gdpPercap),
            gdpPercap_max = max(gdpPercap),
            lifeExp_median = median(lifeExp))
```

For each continent, for each year, calculate the number of countries, and median life expectancy,
```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarise(lifeExp = median(lifeExp),
            gdpPercap = median(gdpPercap),
            pop = sum(as.numeric(pop)))
```

For each year, find the three countries with the lowest GDP per capita,
```{r}
gapminder %>%
  group_by(year) %>%
  arrange(gdpPercap) %>%
  slice(1:3)
```
Note that `arrange()` sorts within the group, not the whole data frame.

For each country, calculate the z-score of its life expectancy,
```{r}
gapminder %>%
  group_by(country) %>%
  mutate(life_exp_z = (lifeExp - mean(lifeExp)) / sd(lifeExp))
```
Note that when a data frame is grouped, summary functions like `mean` and `sd` are calculated within the groups, not over the whole data frame.

For each year, find all observations with greater than the median life expectancy,
```{r}
gapminder %>%
  group_by(year) %>%
  filter(lifeExp > median(lifeExp)) %>%
  arrange(desc(lifeExp))
```

Calculate the difference in log GDP per capita for each country, and the percent growth in GDP (these should be approximately the same for small values),
```{r}
gapminder %>%
  arrange(country, year) %>%
  group_by(country) %>%
  select(country, year, gdpPercap) %>%
  mutate(diff_log_gdpPercap = log(gdpPercap) - lag(log(gdpPercap)),
         gr_gdpPercap = (gdpPercap - lag(gdpPercap)) / lag(gdpPercap))
```

# Plotting

It can be useful to gather the data so that there is only one column with values in order to plot multiple variables or to summarize.

```{r}
gapminder_by_variable <-
  gapminder %>%
  gather(variable, value, -country, -year, -continent)
gapminder_by_variable
```

Now we can plot histograms of all the variables in a single plot,
```{r message = FALSE}
ggplot(gapminder_by_variable, 
       aes(x = value)) +
  geom_histogram() +
  facet_wrap(~variable, ncol = 1, scales = "free")
```

With a column of variables, we can group by that variable, and easily calculate averages,
```{r}
gapminder_by_variable %>%
  group_by(variable, country) %>%
  summarise(mean = mean(value),
            median = median(value),
            max = max(value))
```
Or calculate averages by country, and then transform back to the original form of the variable,
```{r}
gapminder_by_variable %>%
  group_by(variable, country) %>%
  summarise(value = mean(value)) %>%
  spread(variable, value)
```
Create a new data frame in which each row is a country, and columns are the means and standard deviations of `gdpPercap`, `pop`, and `lifeExp` per country,
```{r}
gapminder_by_variable %>%
  group_by(variable, country) %>%
  summarise(mean = mean(value),
            sd = sd(value)) %>%
  gather(stat, value, -country, -variable) %>%
  unite(varstat, variable, stat) %>%
  spread(varstat, value)
```


