---
title: Reading Data in R
date: January 9, 2016
author: Andreu Casas
---

```{r init,results='hide',echo=FALSE}
source("init.R")
```

## Learning Objectives

1. Reading data form R packages
2. Read and write csv files
3. Read and write other type of data formats: STATA, SPSS, XLS...
4. Replication, Dataverse, and getting data from an external source

## R Projects

Keeping all the files associated with a project organized together – input data, R scripts, analytic results, figures – is such a wise and common practice that RStudio has built-in support for this via its projects.  Read [this](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects) for more information about RStudio projects.

You will use RStudio projects for your labs and homework. Create a RStudio project that you will use for all your labs.

1. *File -> New Project*
2. Select "New Directory"
3. Select "Empty Project"
4. Select a name for your project as Directory Name.
   "POLS_501_Labs" is as good as any, and better than most.
   Then choose where to put this directory with "Create project as sub-directory of".
   Don't worry about the other options.
5. Create a new directory within "POLS_501_Labs" called "lab1"
6. Create a new directory within "lab1" called "data"
7. Open this new project and open a new R script where you will be writing (and copy-pasting) and running code.

In lab I'll do live-coding and we'll use R scripts to take notes and write the code:

1. Open a new R script
2. Write a title 
3. Your name (and email?)
4. Date

## 1. Reading data from R packages

Some R packages have datasets that you can load if you have those packages installed and loaded. The `base` package already comes with some datasets. To view the datasets, use the function `data()`. To load a specific dataset, use the same function and add the name of the dataset as a parameter: e.g. `data(UCBAdmissions)`. 


`r challenge_start()`
    
- Load any dataset you like
- Explore it. Tell me what type of information contains, its dimensions, variables, etc.
- Use the functions:

  - `dim` shows the dimensions of the data frame as the number of rows, columns
  - `str` shows the internal structure of an R object
  - `names` shows the column names of the data frame.
  - `head` shows the first few observations
  - `tail` shows the last few observations
  - `summary` calculates summary statistics for all variables in the data frame.
`r challenge_end()`

When we load new packages that have datasets within them, R adds them to the `data()` list. Actually some packages have no functions, only dataset/s: e.g. `gapminder` and `HistData`. Let's load the `HistData` package and take a look at the datasets it contains using the `data()` function and also the `help` file of the package.


```{r, eval=FALSE, results='hide'}
install.packages("HistData")
```
```{r results='hide'}
library("HistData")
```

```{r eval=FALSE}
data()
?HistData
```

The `help` files give us a brief description of packages, functions, and its parameters. A very useful characteristic of `help` files is that they come with sample code (scroll-down to the bottom of the `help` file to find it).

Let's explore a `help` file: click to the `help` file for the `Snow` dataset in `HistData`. 

`r challenge_start()`
Do you know about John Snow's 1854 London Cholera Outbreak study and his visualization of the Pump on Broad Street? Google it and see if you can find the [map-visualization](http://static.guim.co.uk/sys-images/Guardian/Pix/pictures/2013/3/14/1363276044871/John-Snows-cholera-map-of-009.jpg).
`r challenge_end()`

Now let's try to replicate his map using this R package/dataset. Scroll down to the bottom of the help file and copy and runt the following lines of code:

```{r}
data(Snow.deaths); data(Snow.pumps); data(Snow.streets); data(Snow.polygons)
Sdeaths <- function(col="red", pch=15, cex=0.6) {
  	# make sure that the plot limits include all the other stuff
   plot(Snow.deaths[,c("x","y")], col=col, pch=pch, cex=cex,
   	xlab="", ylab="", xlim=c(3,20), ylim=c(3,20),
   	main="Snow's Cholera Map of London")
	}
Spumps <- function(col="blue", pch=17, cex=1.5)  {
   points(Snow.pumps[,c("x","y")], col=col, pch=pch, cex=cex)
   text(Snow.pumps[,c("x","y")], labels=Snow.pumps$label, pos=1, cex=0.8)
}
Sstreets <- function(col="gray") {
  slist <- split(Snow.streets[,c("x","y")],as.factor(Snow.streets[,"street"]))
	invisible(lapply(slist, lines, col=col))
}
```

Finally, let's plot the dataset using the functions we just created (copied from the help file).

```{r}
Sdeaths()
Spumps()
Sstreets()
```

## Read and write csv files

One of the most common data formats is the CSV (Comma Separated Values). CSV datasets are easy to share and read using multiple statistical software. 

Download the dataset **wdi_sample.csv** from the following [link](https://GitHub.com/UW-POLS501/pols_501_wi16/blob/master/lessons/data/wdi_subset.csv) and save it in `lab1/data`. To read in the CSV file, use the `read.csv()` function, which has the main following parameters:

- `file` name of the file (e.g. 'data/wdi_sample.csv')
- `header` (TRUE/FALSE) whether the first row contains variable names (TRUE default value)
- `sep` character that separates variables (',' default value)

```{r}
wdi <- read.csv(file = "data/wdi_sample.csv")
```

Briefly explore the dataset

```{r,results='hide'}
dim(wdi)
head(wdi)
names(wdi)
unique(wdi$Country.Name)
unique(wdi$Indicator.Name)
wdi[50, ]
```

`r challenge_start()`
What are the values of the `country` and `wdi indicator` in the first row of the dataset? What are the values of `country` and `indicator` in the last row?

```{r, results='hide'}
head(wdi$Country.Name)
head(wdi$Indicator.Name)
tail(wdi$Country.Name)
tail(wdi$Indicator.Name)
```
`r challenge_end()`

CSV files can easily be manipulated using a Spreadsheet editor such as Excel. Let's open the `wdi_sample.csv` in Excel (or another editor) and:

- Add an extra row
- Save the file
- Read in the CSV file in R again (remove first the `wdi` dataset that we previously loaded using the function `rm()`)

Explore the dataset to see if the observation we added is there. Add an extra column/variable to the dataset.


```{r}
wdi$test_var <- 1
```

Save the new version of the dataset in CSV format

```{r}
write.csv(x = wdi, file = "data/wdi_sample2.csv")
```



## Read and write other type of data formats: STATA, SPSS, XLS, XLSX...

Not all datasets we want to use are in CSV format. They are often in other formats such as `.txt`, `.sav` (SPSS's native format), `.dta` (STATA), `.xls` or `.xlsx` (Excel formats), etc. There are some R packages that make it easy to read those data formats into R. One of them is the ``rio`` package. Download the following datasets, install the ``rio`` package, and load them into R using ``rio``

- [phd.txt](https://GitHub.com/UW-POLS501/pols_501_wi16/blob/master/lessons/data/phd.txt) Time to PhD dataset from Espenshade and Rodriguez (1997) SSQ, available [here](http://data.princeton.edu/wws509/datasets/#phd).

- [edu_expend_1975.sav](https://GitHub.com/UW-POLS501/pols_501_wi16/blob/master/lessons/data/edu_expend_1975.sav) Education Expenditure 1960, from Chatterjee, Hadi and Price (1977), available [here](http://www.ats.ucla.edu/stat/spss/examples/chp/chpspss_dl.htm)

- [salary.dta](https://GitHub.com/UW-POLS501/pols_501_wi16/blob/master/lessons/data/salary.dta) Discrimination in Salaries, from Weisber (1985), available [here](http://data.princeton.edu/wws509/datasets/#salary)

- [divorce.xlsx](https://GitHub.com/UW-POLS501/pols_501_wi16/blob/master/lessons/data/divorce.xlsx) Marriage Dissolution in the U.S., from Lillard and Panis (2000), available [here](http://data.princeton.edu/wws509/datasets/#divorce)

```{r eval=FALSE,results="hide"}
install.packages("rio")
```
```{r results="hide"}
library("rio")
phd <- import("data/phd.txt")
edu <- import("data/edu_expend_1975.sav")
salary <- import("data/salary.dta")
divorce <- import("data/divorce.xlsx")
```

`r challenge_start()`
Explore the datasets that we just loaded using some of the functions that we have already seen. Check if they all have been correctly loaded.
`r challenge_end()`

We observe 2 issues when reading in the `phd` dataset: it has no variable names in the first row, and R believes the datset has only 1 variable. Let's read the dataset in using the `read.table()` function of the `utils` package

```{r}
phd <- read.table("data/phd.txt")
```

The `rio` package is very useful but not perfect. The following are other functions and packages you may consider when importing datasets into R:

- `excel_sheets` and `read_excel` from the `readxl` package: to import excel files
- `foreign` package has multiple functions to import numerous data formats such as `.sav`, `.dta`, etc.




## Replication, Dataverse, and getting data from an external source

Replication and transparency are key components of all scientific research. However, in the past social scientists have often not been very transparent. For this reason, there are currently numerous initiatives aiming to increase transparency in social science research. One of the main objectives of this course is to learn how to produce clear guidelines when developing our research so others can easily replicate it in the future. 

As part of these replication efforts, authors and journals often share replication datasets and code in their websites or online repositories. One of the most used repositories is [`Dataverse`](https://dataverse.harvard.edu/). For example, the American Journal of Political Science (AJPS) posts replication files in [dataverse](https://dataverse.harvard.edu/dataverse/ajps) for all the articles published in the journal. 

For example, AJPS recently posted the replication files of an article by Broockman and Bulter (2015) 'The Causal Effects of Elite Position-Taking on Voter Attitudes: Field Experiments with Elite Communication'. In this article the authors perform two experiments to show how opinions stated by public representatives have the potential to change the opinions of their constituents. You can access the replication files [here](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/YZQTAH) and the paper [here](http://onlinelibrary.wiley.com/doi/10.1111/ajps.12243/abstract). 

`r challenge_start()`
Go to the replication site of the paper and download the replication dataset `Study1_data.tab`. You can choose the format of the dataset, choose to download it in `RData` format. Save it in the working the directory you are using for this lab. 

Now load it and play with it! `.Rdata` is R's native format. To load an `.Rdata` (or `.rda`) dataset, you use the `load()` function and you don't need to assign the dataset to any object, the `load()` function will automatically create a new dataset in your R environment (similar to what happens when we use the `data()` function to load datasets from R packages).

```{r, results='hide'}
load("data/Study1_data.Rdata")
```
`r challenge_end()`

`r challenge_start()`
Look at the article and explore the dataset and see if you can identify in the dataset some of the key variables they discuss in the paper. 

```{r, results='hide'}
names(x)
table(x$policy_letter_treat)
table(x$policy_letter_treat,x$movable_total)
```
`r challenge_end()`

