---
title: "For loops"
author: "Jeffrey Arnold"
date: "February 4, 2016"
---

```{r results='hide',echo=FALSE}
source("init.R")
```


## For Loops

Suppose we want to print each word in a sentence.
One way is to use six `print` statements:

```{r}
best_practice <- c("Let", "the", "computer", "do", "the", "work")
print(best_practice[1])
print(best_practice[2])
print(best_practice[3])
print(best_practice[4])
print(best_practice[5])
print(best_practice[6])
```

but that's a bad approach for two reasons:

 1. It doesn't scale: if we want to print the elements in a vector that's hundreds long, we'd be better off just typing them in.

 2. It's fragile: if we give it a longer vector, it only prints part of the data, and if we give it a shorter input, it returns `NA` values because we're asking for elements that don't exist!

```{r}
sentence[-6]
```

Here's a better approach:

```{r}
for (something in best_practice) {
  print(paste("this word is ", something))
}
```

This is shorter---certainly shorter than something that prints every character in a hundred-letter string---and more robust as well.
If you change what is in sentence, it will still work.
```{r}
best_practice <- c("a", "b")
for (word in best_practice) {
  print(word)
}
```

The improved version of the code uses a *for loop* to repeat an operation---in this case, printing---once for each thing in a collection.
The general form of a loop is:

```r
for (variable in collection) {
  do things with variable
}
```

We can name the loop variable anything we like (with a few restrictions, e.g. the name of the variable cannot start with a digit).
`in` is part of the `for` syntax.
Note that the body of the loop is enclosed in curly braces `{ }`.
For a single-line loop body, as here, the braces aren't needed, but it is good practice to include them as we did.

[restrictions]: http://cran.r-project.org/doc/manuals/R-intro.html#R-commands_003b-case-sensitivity-etc

Here's another loop that repeatedly updates a variable:

```{r}
len <- 0
vowels <- c("a", "e", "i", "o", "u")
for (v in vowels) {
  len <- len + 1
}
# Number of vowels
len
```

It's worth tracing the execution of this little program step by step.
Since there are five elements in the vector `vowels`, the statement inside the loop will be executed five times.
The first time around, `len` is zero (the value assigned to it on line 1) and `v` is `"a"`.
The statement adds 1 to the old value of `len`, producing 1, and updates `len` to refer to that new value.
The next time around, `v` is `"e"` and `len` is 1, so `len` is updated to be 2.
After three more updates, `len` is 5; since there is nothing left in the vector `vowels` for R to process, the loop finishes.

Note that a loop variable is just a variable that's being used to record progress in a loop.
It still exists after the loop is over, and we can re-use variables previously defined as loop variables as well:

```{r}
letter <- "z"
for (letter in c("a", "b", "c")) {
  print(letter)
}
# after the loop, letter is
letter
```

Note also that finding the length of a vector is such a common operation that R actually has a built-in function to do it called `length`:

```{r}
length(vowels)
```

`length` is much faster than any R function we could write ourselves, and much easier to read than a two-line loop; it will also give us the length of many other things that we haven't met yet, so we should always use it when we can

We can name the loop variable anything we like (with a few restrictions, e.g. the name of the variable cannot start with a digit).
`in` is part of the `for` syntax.
Note that the body of the loop is enclosed in curly braces `{ }`.
For a single-line loop body, as here, the braces aren't needed, but it is good practice to include them as we did.


`r challenge_start()`
R has a built-in function called `seq` that creates a list of numbers:

```{r}
seq(3)
```

Using `seq` use a for loop to print the first **N** natural numbers, one per line:
`r challenge_end()`
`r solution_start()`
```{r, echo=-1}
N <- 3
nseq <- seq(N)
for (num in nseq) {
  print(num)
}
```
`r solution_end()`

Building a data frame from a list of data frames.

```{r}
means <- c(-5, 0, 4)
n <- 5
results <- list()
for (i in 1:length(means)) {
  results[[i]] <-
    data.frame(x = rnorm(n, mean = means[i]),
               mu = means[i])
}
```
The object `results` is a list of data frames,
```{r}
results
```
We can stack these data frames into a single data frame using the **dplyr** function `bind_rows`,
```{r}
library("dplyr")
bind_rows(results)
```



* * * 

This is derived from the  Software Carpentry lesson, [Programming in R](http://swcarpentry.github.io/r-novice-inflammation/), [CC-BY](http://software-carpentry.org/license/).
