---
title: "Missing Values in R"
author: Jeffrey Arnold
date: 2016-01-21
output: html_document
---

```{r results='hide',echo=FALSE}
source("init.R")
```

In almost any set of real data, there will be missing values.
All statistical software systems have some way of indicating missing values within variables.
R uses the symbol `NA` to indicate a missing value. 

# Behavior of NA

Create two vectors that include missing values,
```{r}
foo <- c(1, NA, 3)
bar <- c(4, 5, NA)
```

`r challenge_start()`

Try the following commands:. How do the NA affect the results?

```{r ch1,eval = FALSE}
foo + bar
foo - bar
foo * bar
foo / bar
foo == 1
foo > 1
sum(foo)
max(foo)
```

`r challenge_end()`

`r solution_start()`

For the arithmetic operators, `+-*/`, any element with an `NA` produces an `NA` in the output.
For the logical operators, comparing an element with `NA` also produces an `NA` (not `FALSE` as might be expected).
For `sum` and `mean`, any `NA` in the vector produces `NA` for a result.

```{r}
<<ch1>>
```

`r solution_end()`

# Testing for `NA` with `is.na`

Since comparing any value to `NA` produces an `NA`, you can't easily use `==` to test whether an element of a vector is equal to `NA`. Why not? 

The function `is.na` is used to test whether an element of a vector is equal to `NA`.
```{r}
is.na(foo)
is.na(bar)
```
Thus, to test whether an element of a vector is *not* equal to `NA`,
```{r}
!is.na(foo)
!is.na(bar)
```

`r challenge_start()`
Using `is.na` and `[` drop all missing values from `foo`.
`r challenge_end()`

`r solution_start()`
```{r}
foo[!is.na(foo)]
```

To check whether there are any missing values in a vector,
```{r}
any(is.na(foo))
```
To check whether all values in a vector are missing,
```{r}
all(is.na(foo))
```
To count the number of missing values in a vector,
```{r}
sum(is.na(foo))
```
or calculate the fraction of missing values,
```{r}
sum(is.na(foo)) / length(foo)
```


`r solution_end()`

# Replace missing values

There are several ways to replace missing values in a vector,
```{r}
alpha <- c(1, 2, NA)
alpha[is.na(alpha)] <- 0
alpha
```
or
```{r}
beta <- c(1, 2, NA)
beta <- ifelse(is.na(beta), 0, beta)
```

# Set values to missing

Sometimes you need to set common

A common convention in old datasets and survey data is to indicate missing values using
unusually large or small values. For example, -66, -77, -88, or -99 in a 0-10 scale.
How would you replace them with missing values?

Consider the following vector.
```{r}
Q1 <- c(3, -88, 8, 2, 6, 7, 1, 5, 1, 9, -99)
```
Suppose, it represents the response to a question on a 0-10 scale, -88 is "Don't know", and
-99 is "Refused to answer".
For your analysis, it makes sense to to "Don't know" and "Refused to answer"" with missing values.

There are a few ways to do this: using `[`, using `is.na`, and using `ifelse`.

```{r}
Q1_clean1 <- ifelse(Q1 %in% c(-88, -99), NA, Q1)
Q1_clean1

Q1_clean2 <- Q1
Q1_clean2[Q1_clean2 %in% c(-88, -99)] <- NA
Q1_clean2

Q1_clean3 <- Q1
is.na(Q1_clean3) <- Q1_clean3 %in% c(-88, -99)
Q1_clean3
```


# Missing values in character vectors

`NA` values can also occur in character vectors,
```{r}
baz <- c("A", "B", NA, "")
```
In a character vector, `NA` is **not** the same as "".

`r challenge_start()`

Start with the vector of capital Latin letters.
```{r}
letr <- LETTERS
```
Note that the variables `letters` and `LETTERS`
have the 26 Latin alphabet letters in lowercase and uppercase, respectively.

Randomly set 3 elements to NA, and 3 elements to ""
```{r}
letr[sample(1:length(letr), 6)] <- c(rep(NA, 3), rep("", 3))
```
Without explicitly referencing the indexes of the elements

1. create a new vector named `letr` with all `NA` values to "".
2. create a new vector with all "" equal to `NA`
3. Remove all elements that are either "" or `NA`

`r challenge_end()`

`r solution_start()`
To replace `NA` with "",
```{r}
letr1 <- letr
letr1[is.na(letr1)] <- ""
letr1
```
or using `ifelse`,
```{r}
letr2 <- ifelse(is.na(letr), "", letr)
letr2
```
or using `is.na`,
```{r}
letr2 <- letr
is.na(letr2) <- letr2 == ""
```

To replace "" with `NA`,
```{r}
letr2 <- ifelse(letr == "", NA, letr)
letr2
```

To drop all observations with `NA` or "",
```
letr3 <- letr[! letr %in% c(NA, "")]
letr3
```
`r solution_end()`


# Reading text files (csv)

The function `read_csv` has an argument `na.strings` which determines what cell values are converted to `NA`. By default it only reads cells with "NA" as missing values.
However, often you will want to treat any blank cells, e.g. `""` as missing. 
To do that,
```{r eval=FALSE}
read.csv(file, na.strings = "")
```


# Other functions

**tidyr** has a function `replace_na` which fills in missing values in multiple columns of a data frame,
```{r}
library("tidyr")
library("dplyr")
data_frame(x = c(1, 2, NA), y = c("a", NA, "b")) %>%
  replace_na(list(x = 0, y = "unknown"))
```

**stringr** has a function `str_replace_na` to replace `NA` values in a character vector,
```{r}
library("stringr")
str_replace_na(c("a", "b", NA), "")
```

# Listwise deletion of rows with missing values in data frames

To drop any row in a data frame with any missing values use `na.omit`,
```{r}
na.omit(data.frame(a = c(NA, 1, 2, 3), b = c("a", "b", NA, "d")))
```


# Imputing or filling missing values

The package *zoo* contains several functions to fill in missing values in time series data.
`na.locf` fills in missing values with the last non-missing value,
```{r}
library("zoo")
na.locf(c(1, NA, NA, 2, NA, NA, NA, 3, 4, 5))
```
`na.approx` fills in missing values with a linear interpolation, while `na.spline` fills in missing observations with a spline (smooth) interpolation,
```{r}
plot(na.spline(c(1, NA, NA, 2, NA, 4, NA, NA, NA, NA, 3, NA, 1)), type = "l")
plot(na.approx(c(1, NA, NA, 2, NA, 4, NA, NA, NA, NA, 3, NA, 1)), type = "l")
```


See packages:

- [mice](https://cran.r-project.org/web/packages/mice/index.html)
- [Amelia](https://cran.r-project.org/web/packages/Amelia/index.html)



- http://www.cookbook-r.com/Manipulating_data/

<!--

- Talk about NULL? 
- Don't need to mention Inf or NaN, since those appear infrequently in what they are doing

-->
