---
title: "The Pipe Operator"
author: "Jeffrey Arnold"
date: "January 21, 2016"
output: html_document
---

```{r result='hide',echo=FALSE}
source("init.R")
knitr::opts_chunk$set(results = 'hide')
```

```{r}
library("dplyr")
library("magrittr")
```

The symbol `%>%` is known as the pipe operator. It is from the **magrittr** package,
but most widely used in the **dplyr** package. 

The verbal description of what the pipe operator does is it takes the result of the previous expression and uses it as the first argument in the next function.

For example, suppose we wanted to take a sample of 100 from the normal distribution and find the maximum value in the sample.
```{r}
n <- 100
max(rnorm(n))
```
Using the `%>%` we can write this as,
```{r}
rnorm(n) %>% max()
```
or, even,
```{r}
n %>% rnorm() %>% max()
```

The pipe operator is particularly useful with **dplyr**.

Let's use the `gampinder` data, and find the average life expectancy of the median country in each continent in 2007 using **dplyr**,
```{r}
library("gapminder")
data("gapminder")
summarise(group_by(filter(gapminder, year == 2007),
                   continent),
          lifeExp = median(lifeExp))
```
We could also write it while storing the results,
```{r}
tmpdata <- filter(gapminder, year == 2007)
tmpdata <- group_by(tmpdata, continent)
tmpdata <- summarise(tmpdata, lifeExp = median(lifeExp))
tmpdata
```
Now compare those to how we would write it using `%>%`,
```{r}
gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarise(lifeExp = median(lifeExp))
```

`r challenge_start()`

Rewrite this code to use `%>%`:

```{r results='hide'}
summarize(group_by(gapminder, continent),
          n_obs = n(), n_countries = n_distinct(country))


tmp <- group_by(gapminder, continent, country)
tmp <- select(tmp, country, year, continent, lifeExp)
tmp <- mutate(tmp, le_delta = lifeExp - lag(lifeExp))
tmp <- summarize(tmp, worst_le_delta = min(le_delta, na.rm = TRUE))
tmp <- filter(tmp, min_rank(worst_le_delta) < 2)
tmp <- arrange(tmp, worst_le_delta)
tmp
```
`r challenge_end()`
`r solution_start()`
```{r}
gapminder %>%
  group_by(continent) %>%
  summarize(n_obs = n(), n_countries = n_distinct(country))

gapminder %>%
  group_by(continent, country) %>%
  select(country, year, continent, lifeExp) %>%
  mutate(le_delta = lifeExp - lag(lifeExp)) %>%
  summarize(worst_le_delta = min(le_delta, na.rm = TRUE)) %>%
  filter(min_rank(worst_le_delta) < 2) %>%
  arrange(worst_le_delta)

```
`r solution_end()`



