---
title: "Probability and Distributions in R"
date: January 29, 2016
author: Andreu Casas
---

```{r init,results='hide',echo=FALSE}
source("init.R")
```

## Learning Objectives
1. What is uncertainty and types of uncertainty
2. What are probabilities and different types of distributions
3. How to use them in R

## Sampling, Theoretical, and Fundamental Uncertainty

**What is sampling uncertainty?**

SMISS (p.73): "Uncertainty about the characteristics of a whole that remains after observing a part"

We will use the **gapminder** dataset to better understand *sampling uncertainty*:

- Subset only the observations in **gapminder** from 2007 (imagine this is the population: all existing countries)
- Calculate the mean and standard deviation of **lifeExp**
- Subset 10 random countries from the 2007 dataset
- Calculate the mean and standard deviation of **lifeExp** for those 10 obs.

```{r results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(gapminder)
data(gapminder)
set.seed(123)
```

```{r, results = 'hide'}
data2007 <- gapminder %>%
  filter(year == "2007")

x <- mean(data2007$lifeExp) 
s <- sd(data2007$lifeExp)

random_10_countries <- sample(gapminder$country, 10)
data2007_10 <- data2007 %>%
  filter(country %in% random_10_countries)

x_hat <- mean(data2007_10$lifeExp)
s_10 <- sd(data2007_10$lifeExp)
```

For a particular case (since we randomly sample 10 countries, each time we repeat the analysis we get slightly different results) we obtain that:

```{r xtable, results='asis', echo=FALSE}
library(xtable)
table1 <- cbind(x , s)
table1 <- rbind(table1, cbind(x_hat, s_10))
colnames(table1) <- c('Mean', 'SD')
rownames(table1) <- c('Population','Sample')
print(xtable(table1), type='html')
```

\n 

`r challenge_start()`
What are the *theoretical* and *fundamental* uncertainties?
`r challenge_end()`

## Probability and Distributions in R

How many heads and how many tails would we get if we flip a fair coin 10 times?
 
```{r}
flips_10 <- sample(x = c('head','tails'), size = 10, replace = TRUE)
table(flips_10)
```

According to the last (Jan. 26, 2016) *polls-only* forecast of [FiveThirtyEight](http://projects.fivethirtyeight.com/election-2016/primary-forecast/iowa-republican/#polls-only), Donald Trump has a 55% probability of winning the Iowa caucuses. 

`r challenge_start()`
How can we simulate the Iowa caucuses to make one prediction of Trumps's victory/loss? What type of distribution represents this outcome variable? 
`r challenge_end()`

```{r}
# Bernoulli Distribution: a Bernoulli Trial
trump_result <- rbinom(size = 1, # Number of trials
                       n = 1,    # Number of observations
                       p = .55)  # Probability of succes on each trial
print(c("Trump Loses :(", "Trump Wins! :)")[trump_result + 1])
```

`r challenge_start()`
How can we simulate Trump's victory or loss in Iowa 10 times using R? What type of distribution represents this outcome variable? 
`r challenge_end()`

```{r}
# Binomial Distribution: N draws from a Bernoulli distribution
trump_result_10 <- rbinom(size = 10, # Number of trials/draws
                       n = 1,    # Number of observations (e.g. 0-1)
                       p = .55)  # Probability of succes on each trial
print(paste("Trump wins the election in",trump_result_10,"of the simulations"))
```


Now we will use the **country.csv** dataset we used in previous labs. 

```{r}
library(ggplot2)
country <- read.csv("data/country.csv")
```

Remember that this is a simplified dataset for the Baum and Zhukov's (2015) *JPR* article where the variable **n_articles** is the number of news stories that newspapers from these countries dedicated to talk about the Lybian Civil Crisis.

Lets plot the distribution of the variable **n_articles** using a density plot.
```{r}
ggplot(country, aes(x = n_articles)) + 
  geom_density()
```

`r challenge_start()`
What type of distribution does the variable **n_articles** have? (Hint: distribution for a variable of nonnegative integers)
`r challenge_end()`

*Poisson Distribution:* Poisson($\lambda$) = E(Y) = Var(Y) = $\lambda$

Let's plot a Probability Mass Function (PMF) of a Poisson distribution with the mean of **n_articles**. Do you think this PMF is similar to the distribution of **n_articles** we previously plot?

```{r}
n_articles_mean <- mean(country$n_articles)
y_articles <- 1:max(country$n_articles)
pmf <- dpois(y_articles, lambda = n_articles_mean) 
pmf <- data.frame(pmf)
ggplot(pmf, aes(x = y_articles, y = pmf)) + geom_line()
```




