---
title: "reordering-labels-plot"
date: January 29, 2016
author: Andreu Casas
---

```{r init,results='hide',echo=FALSE}
source("init.R")
```

## Learning Objectives
How to sort the values ploted in a graph by a specific variable

## Sorting Labels of a Plot
We start by creating a simulated dataset with data about:

- `parties` Name of political parties
- `ideology` An ideology score for that party
- `women_per` The percentage of party candidates that are women

```{r}
parties <- paste0("party",1:10)
ideology <- sample(x = 1:10, size = 10, replace = FALSE)
women_per <- rnorm(n = 10, mean = 0.5, sd = 0.2)
dataset <- data.frame(parties = parties, ideology = ideology, women_per = women_per)
```

We use `ggplot2` to create a barplot showing the percentage of women candidate by political party. However, in the x-axis, the parties are sorted alphabetically. 
```{r}
library(ggplot2)
ggplot(dataset, aes(x = parties, y = women_per)) + 
  geom_bar(stat = "summary", position = "dodge", fun.y="sum")
```

If we want to sort them by another variable (e.g. `ideology`), we need to specify so in the ggplot comand using `reorder()`.
```{r}
ggplot(dataset, aes(x = reorder(parties,ideology), y = women_per)) + 
  geom_bar(stat = "summary", position = "dodge", fun.y="sum")
```

Make sure the variable you are using to sort the axis is numeric, otherwise you will get an error and it won't work.
```{r, eval = FALSE}
dataset$ideology <- as.factor(dataset$ideology)
ggplot(dataset, aes(x = reorder(parties,ideology), y = women_per)) + 
  geom_bar(stat = "summary", position = "dodge", fun.y="sum")
```

If it's not a numerical variable, change it so it is. 
```{r}
dataset$ideo_numeric <- as.numeric(dataset$ideology)
ggplot(dataset, aes(x = reorder(parties,ideo_numeric), y = women_per)) + 
  geom_bar(stat = "summary", position = "dodge", fun.y="sum")
```



