---
title: "Student’s t-distribution and Inference of the Population Mean"
author: "Jeffrey Arnold"
date: "February 10, 2016"
output: html_document
---

# Sampling Distribution

Asymptotically, the Central Limit Theorem states that the sampling distribution of the sample 
mean is $N(\mu, \sigma / \sqrt{n})$.
This means that test-statistics and critical values are defined as,
$$
\frac{\bar{x} - \mu}{\sigma / \sqrt{n}} \sim N(0, 1)
$$
The problem with this is that rarely do we know the population standard deviation, $\sigma$.
It would be unusual to have an unknown population mean, but a known population standard deviation.
So in practice, the test statistic that is used plugs in the sample standard deviation for the population standard deviation
$$
\frac{\bar{x} - \mu}{\sigma / \sqrt{n}} \approx \frac{\bar{x} - \mu}{s / \sqrt{n}}
$$
But the sample variance, and by extension, the sample standard deviation, will vary across samples,
```{r}
for (i in 1:5) {
  var(rnorm(10))
}
```
So the sampling distribution of the test statistic of the mean will need to account for both the variability of sample means and standard deviations.

The sample standard deviation has its own sampling distribution.
Let's draw samples from the standard normal distribution at several sample sizes, and save the mean and variance of each sample.
```{r}
library("dplyr")
samples <-
    data_frame(size = as.integer(c(4, 16, 32, 256)),
               iter = 2048) %>%
    group_by(size) %>%
    do({
      results <- list()
      for (i in 1:.$iter) {
        x <- rnorm(.$size)
        results[[i]] <- data_frame(x_mean = mean(x),
                                   x_var = var(x))
      }
      bind_rows(results)
    }) %>% ungroup()
```

```{r}
ggplot(samples, aes(x = x_var, colour = factor(size))) +
  geom_density()
```

The sampling distribution of the variance is known when the population distribution is a normal distribution. 
In that case, the sampling distribution for the sum of squares of iid samples 
$$
(n - 1) s^2  / \sigma^2 \sim \chi^2_{n - 1}
$$
```{r}
samples_w_chisq <- samples %>%
  mutate(sum_sq = x_var * (size - 1),
         chisq = dchisq(x_var * size, df = size),
         size = factor(size))
ggplot(samples_w_chisq) +
  geom_density(mapping = aes(x = sum_sq), color = "black") +
  geom_line(mapping = aes(x = sum_sq, y = chisq), color = "red") +
  facet_wrap(~size, scales = "free_x")
```

For smaller sample sizes the tails of the distribution of $(\bar{x} - \mu) / (s / \sqrt{n})$ are wider than those of $(\bar{x} - \mu) / (\sigma / \sqrt{n})$. 
This is due to the added sampling variability of the sample variance (standard deviation).
```{r}
comparison <- samples %>%
  mutate(known = x_mean / (1 / sqrt(size)),
         estimated = x_mean / sqrt(x_var / size)) %>%
  select(size, known, estimated) %>%
  gather(sigma, value, -size)
ggplot(comparison, aes(samples = value, colour = sigma)) +
  geom_qq() +
  facet_wrap(~size, scales = "free")
```


The Student's t distribution is the distribution of the sampling distribution of the mean with an unknown standard deviation.


TODO: coverage ratios of confidence intervals

TODO: Type I errors for sample sizes with normal and student's t distribution


