---
title: Subsetting and Plotting Data in R
date: January 15, 2016
author: Andreu Casas
---


```{r echo=FALSE,results='hide'}
source("init.R")
```

### Learning Objectives

1. Improve subsetting skills
2. Plot data using `ggplot2`

***

<center> <h3> Initial Set up </h3> </center>

- Create new R project for this lab
- Create a **data** subdirectory

Today we will work with data from a recent [article](http://jpr.sagepub.com/content/52/3/384.full):

- Baum, Matthew A. and Yuri M. Zhukov. (2015) "Filtering revolution: Reporting bias in international newspaper coverage of the Libyan civil war." *Journal of Peace Research*, 52(3):384-400.

This article won the [2015 Best Visualization Award](https://www.prio.org/JPR/VisualizationAward/), given by the Journal of Peace Research to the article published in the journal that "not only impresses by its interesting analysis, but also by the large number of aesthetically pleasing and informative graphic."

The replication data for the article is available in JPR's [website](https://www.prio.org/JPR/Datasets/). However, since the replication material for this paper takes a while to download, and the four datasets they use are large and have numerous variables, for the purpose of this class we will use a version of one of their datasets that I put together. 

- Download the dataset in this [link](https://github.com/UW-POLS501/pols_501_wi16/blob/master/labs/subsetting-and-plotting/data/country.csv) and store it in the **data** subdirectory

**Challenge:**

- Go to the article link, skim it (read the abstract and the data section), and briefly tell me:

    - What the paper is about
    - The type of data they use
    - What they do with the data (how they use it to test their hypotheses)
  
***

<center> <h3> Improve subsetting skills </h1> </center>

#### Selecting variables of our dataset

Start by reading the dataset (a CSV file) into `R`. 

**Challenge:**

- How do we read CSV files into R?

```{r}
country <- read.csv('data/country.csv')
```

Explore the dataset using the functions we learned last week:

- `dim()`
- `str()`
- `names()`
- `head()`
- `tail()`
- `summary()`

**Challenge,** how would we select:

- first 10 rows
- last 10 rows
- second variable (`polity2`)
- first 10 observations of the second variable (`polity2`)
- a subset of the data with only democratic countries
- a subset of the data with only non-democratic countries

```{r, results='hide'}
# First 10 rows
country[1:10, ]
head(country,n = 10)

# Last 10 rows
obs <- nrow(country)
country[(obs-10):obs,]
tail(country, n=10)

# Second variable 'polity2'
country[,2]
country$polity2

# First 10 observations of the second variable ('polity2')
country[1:10 , 2]
country$polity[1:10]
country[,2][1:10]

# A subset of the data with only democratic countries
# See the importance of using MEANINGFUL variable names (we know democratic values have democracy = 1)
table(country$democracy)
demCountries <- country[country$democracy==1, ]
nonDemCountries <- country[country$democracy==0, ]
```



